---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vtamR

<!-- badges: start -->
<!-- badges: end -->

**vtamR** is a revised, completed version of 
[VTAM](https://www.csbj.org/article/S2001-0370(23)00034-X/fulltext) 
(Validation and Taxonomic Assignation of Metabarcoding Data) 
rewritten in R. It is a complete metabarcoding pipeline:

* Sequence analyses **from raw fastq files** of amplicon sequences till 
Amplicon Sequence Variant 
([ASV](https://people.imbe.fr/~emeglecz/vtamR/tutorial-vtamr-pipeline.html#glossary)) 
table of **validated ASVs assigned to taxonomic groups**.
* Handles technical or biological **replicates** of the same sample.
* Uses positive and negative **control samples to fine tune the filtering** and 
reduce [false positive](https://people.imbe.fr/~emeglecz/vtamR/tutorial-vtamr-pipeline.html#glossary) 
and [false negative](https://people.imbe.fr/~emeglecz/vtamR/tutorial-vtamr-pipeline.html#glossary) 
occurrences.
* Can pool multiple data sets (results of earlier analyses)
* Can pool results from overlapping markers

**Novelties compared with VTAM:**

* As a collection of R functions, `vtamR` is highly adaptable, allowing users to
include, exclude, and reorder different analysis steps.
* Includes **SWARM** for denoising, followed by a post-treatment to separate 
abundant ASVs that differ by only one nucleotide but have been clustered together.
* Provides additional **graphical and visualization options**.
* Includes functions to generate statistics for each filtering step (e.g. read and variant counts).
* Generates detailed log files and summaries to facilitate **reproducibility and reporting**, 
including support for the **MIEM checklist**.
* Supports **ASV clustering into mOTUs**.
* Outputs can be easily converted to a **phyloseq** object of data frames ready for **vegan**.
* The concepts of **marker** and **run** have been removed to simplify the analysis workflow.


# Tutorial

A detailed **[Tutorial](https://people.imbe.fr/~emeglecz/vtamR)**
on how to construct a full metabarcoding pipeline is avalibale online.

# Installation


## Implementation

The heavy lifting of sequence analyses are done by the following third party 
programs. They should be installed on your computer:

* BLAST([Altschul et al., 1990](https://pubmed.ncbi.nlm.nih.gov/2231712/)) (v2.11.0+)
* VSEARCH ([Rognes et al., 2016](https://peerj.com/articles/2584/)) (v2.7.0)
* CUTADAPT([Martin, 2011](https://journal.embnet.org/index.php/embnetjournal/article/view/200/479)) (v4.0)
* SWARM ([Mahé et al., 2015](https://peerj.com/articles/1420/)) (v2.1.12)
* PIGZ (optional)

`vtamR` has been tested using the above mentioned versions, 
but it should work with later versions.

`vtamR` was tested on **Windows** and **Linux**, and should work in all 
operating systems.

## Install `vtamR`

Install `vtamR` and its CRAN dependencies using `pak`:

```{r install_github_pak, eval=FALSE}
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
pak::pkg_install("meglecz/vtamR", dependencies = TRUE)
```

**Alternatively**, you can install `vtamR` using `devtools`:

```{r install_github_devtools, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("meglecz/vtamR", dependencies = TRUE)
```

### Optional Bioconductor packages

Some `vtamR` functions require additional Bioconductor packages. These are **suggested packages** and are therefore **not installed automatically with `vtamR`**. Install them manually if you intend to use the corresponding functions:

* `assign_taxonomy_rdp` requires `rRDP`, `rRDPData`, and `Biostrings`.
* `make_phyloseq` requires `phyloseq`.

```{r install_bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

if (!requireNamespace("rRDP", quietly = TRUE)) BiocManager::install("rRDP")
if (!requireNamespace("rRDPData", quietly = TRUE)) BiocManager::install("rRDPData")
if (!requireNamespace("Biostrings", quietly = TRUE)) BiocManager::install("Biostrings")
if (!requireNamespace("phyloseq", quietly = TRUE)) BiocManager::install("phyloseq")
```

## Install third party programs

### Linux

* `VSEARCH` [https://github.com/torognes/vsearch](https://github.com/torognes/vsearch)
* `BLAST` [https://www.ncbi.nlm.nih.gov/books/NBK52640/](https://www.ncbi.nlm.nih.gov/books/NBK52640/)
* `CUTADAPT` [https://cutadapt.readthedocs.io/en/stable/installation.html](https://cutadapt.readthedocs.io/en/stable/installation.html)
* `SWARM` [https://github.com/torognes/swarm](https://github.com/torognes/swarm)
* `PIGZ` (optional) [https://github.com/madler/pigz](https://github.com/madler/pigz)


**Add the executables to your PATH** 

This step is optional. If third-party executables are included in your `PATH`, 
you won’t need to manually specify their locations when running functions that use them.

If you prefer **not** to modify your `PATH`, be sure to provide the 
full path to each program using the corresponding `vtamR` function argument 
(e.g., `vsearch_path`).


1. Edit (or create) your `~/.Renviron` file

```bash
nano ~/.Renviron
```

2. Add the directory to the PATH

```bash
PATH="/home/user/tools/myprogram/bin:${PATH}"
```
**Note:** The PATH must contain a **directory**, not an executable file.
For example, use: `/home/username/miniconda3/envs/vtamr/bin`
and **not**: `/home/username/miniconda3/envs/vtamr/bin/cutadapt`

3. Restart R or RStudio

The new PATH is loaded at startup.

4. Verify that R sees the correct executables

```r
system("cutadapt --version")
system("swarm --version")
system("vsearch --version")
system("blastn -version")
system("pigz --version")
```

### Windows

Download binaries and save them to a convenient place on your computer 
(path without space, e.g. `C:/Users/Public/`)

**VSEARCH**

* Download binaries from https://github.com/torognes/vsearch/releases/tag/v2.23.0
* Decompress the zip file
* The executable are found in `vsearch-x.xx.x-win-x86_64/bin/`

**CUTADAPT**

* Download `cutadapt.exe` from 
[https://github.com/marcelm/cutadapt/releases](https://github.com/marcelm/cutadapt/releases)

**SWARM**

* Download binaries from 
[https://github.com/torognes/swarm/releases](https://github.com/torognes/swarm/releases)
* Decompress the zip file
* The executable is in the `swarm-x.x.x-win-x86_64/bin` directory

**BLAST**

* Download the installer (ncbi-blast-x.xx.x+-win64.exe) from:
https://ftp.ncbi.nlm.nih.gov/blast/executables/LATEST/
* Double-click the .exe file, accept the license agreement, 
and keep the suggested installation path.
This will automatically add BLAST to your system PATH, 
so you won’t need to specify its location when using it.
* The executables will be installed in:
`C:\Program Files\NCBI\blast-x.xx.x+\bin`

**VSEARCH** (Optional)

* Download [pigz-win32.zip](https://sourceforge.net/projects/pigz-for-windows/files/)
* Decompress the zip file
* The executable (pigz.exe) is in the pigz-win32 directory

**Add executables to your PATH**

This step is optional. If third-party executables are included in your 
system PATH, you won’t need to manually specify their locations when running 
functions that use them.

If you prefer **not** to add them, make sure to provide the full path to each program through the appropriate `vtamR` function argument (e.g., `vsearch_path`).

You can use either **GUI** or **PowerShell** to add executables to your PATH.

1. Using the Windows GUI (System Settings)

* Press **Win + R**, type `sysdm.cpl`, and press **Enter**.
   *(Opens System Properties.)*
* Go to the **Advanced** tab.
* Click **Environment Variables…**
* In the **System variables** section, find and select **Path**.
* Click **Edit…**
* Click **New**, then paste the directory you want to add (for example:
   `C:\Users\YourName\Tools\bin`)
* Click **OK** on all windows to save.

2. Using PowerShell

* Press **Win + R**, type `powershell`, and press **Enter**.
   *(Opens Powershell)*

```{}
setx PATH "$($env:Path);C:\My\New\Folder\vsearch:C:\My\New\Folder\cutadapt;C:\My\New\Folder\pigz;C:\My\New\Folder\swarm"
```

* Close de Powershell, since the change applies only to **new** terminals, not the current one.


**Verify Program Versions**

Open a Command Prompt on Windows (**Win + R**, type `cmd`)

```bash
cutadapt --version
swarm --version
vsearch --version
pigz --version
blastn -version
```

**Note**
*These commands will display the version of each program **if the executables are in your PATH**.
Otherwise, don’t forget to provide the full path to the executables.*

Example:

```{}
C:/Users/Public/vsearch-x.xx.x-win-x86_64/bin/vsearch --version
```

## TaxAssign reference data base

### assign_taxonomy_ltg

For `assign_taxonomy_ltg` (BLAST based Lowest Common Ancestor method),
a ready-to-use COI database is available from OSF [OSF](https://osf.io/vrfwz/),
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).

Download and extract using:
```{r, dowload-coinr, eval=FALSE}
library(vtamR)

download_osf(filename = "COInr_for_vtam_2026_05_03_dbV5.tar.gz",
    url = "https://osf.io/download/yhjvm/",
    dest_dir = "COInr_db",
    untar = TRUE,
    quiet = FALSE
    )
```

To check for newer versions:

- Go to: [https://osf.io/vrfwz/files/osfstorage](https://osf.io/vrfwz/files/osfstorage)
- In **Files** open the the **dbV5** folder.
- Open the latest file in the browser (click the file name, not just the folder).
- Look at the URL, it will look like: https://osf.io/yhjvm/ (yhjvm = file ID)
- Build the download link like this: https://osf.io/download/yhjvm/


Alternatively, you can also create a custom version from the **TSV format** database available at
[Zenodo](https://zenodo.org/records/15515860) and make a custom version using 
[mkCOInr](https://github.com/meglecz/mkCOInr)
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).


**For other markers** you will need a [database formatted to BLAST](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments-with-assign_taxonomy_ltg), 
containing taxIDs and a 
[taxonomy file](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments-with-assign_taxonomy_ltg).
If all sequences are extracted from NCBI-nt, the taxonomy file provided in 
[OSF](https://osf.io/vrfwz/) can be used without modification.


### assign_taxonomy_rdp

`assign_taxonomy_rdp` function uses de RDP algorithm to assign bacterial 16S sequences.
The database is provided in the `rRDPData` R data package.

It is also possible to provide a trained classifier object created 
with `trainRDP()`.

