---
title: "vtamR Developer notes"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

# Developper
## Load and roxygenise
```{r}
library("devtools")
library("roxygen2")
setwd("C:\\Users\\emese\\vtamR")
setwd("/home/meglecz/vtamR")
load_all(".")
roxygenise()
usethis::use_roxygen_md()
```


## Creating a package

```{r}
install.packages(c("devtools", "roxygen2", "usethis", "testthat"))
usethis::create_package("path/to/your/packageName")
usethis::use_git()
usethis::use_github()

#### Write the package
devtools::document() # generate doc
# create a template for README.Rmd only the developer Edit. It is in .builtignore
usethis::use_readme_rmd()
# make a README.md from README.Rmd. This should not be edited manually
# When building the readme, the Rmd and its chunks are executed, so all example codes should work. If not, make eval=FALSE
devtools::build_readme()

# creates vignette and adds modif to DESCRIPTION
usethis::use_vignette("vtamR")
# vignettes/ should be edited and built
devtools::build_vignettes()

 # Run checks before releasing a package
devtools::check()     
## OR rcmdcheck 
results <- rcmdcheck::rcmdcheck()
results$errors
results$warnings
results$notes
```

## uninstall
```{r}
remove.packages("vtamR")
```

## create a rda with the asv_tax, so users can load it
```{r}
full_asv_tax <- asv_tax
usethis::use_data(full_asv_tax, overwrite = TRUE)
rm(full_asv_tax)
data(full_asv_tax)
```






## Building new version

- 1. Finalize DESCRIPTION version (e.g. 2.0.0)
- 2. Install package + build vignettes devtools::install(build_vignettes = TRUE)
- 3. Render README.Rmd → README.md devtools::build_readme()
- 4. Rebuild pkgdown site pkgdown::build_site()
- 5. Check git diff on README.md, docs/, vignettes — confirm everything reflects the final version
- 6. Commit everything together, push
- 7. Tag the release (e.g. git tag v2.0.0, or via GitHub UI)

### Update the following files

- DESCRIPTION
- vignettes
- inst/citation
- complete NEWS.md

### load and roxygenise
```{r}
library("devtools")
library("roxygen2")
setwd("C:\\Users\\emese\\vtamR")
setwd("/home/meglecz/vtamR")
load_all(".")
roxygenise()
usethis::use_roxygen_md()
```

### check package
```{r}
usethis::use_roxygen_md()
library("devtools")
devtools::check()  
```

### install package localy and build vignettes (or not)
```{r}
devtools::install(build_vignettes = TRUE)
devtools::install(build_vignettes = FALSE)
```

### build_readme if README.rm has been changed
```{r}
devtools::build_readme()
```

### If pb with vignette not building correctly
```{r}
find.package("vtamR")
remove.packages("vtamR")
# clean package
unlink("C:/Users/emese/AppData/Local/R/win-library/4.4/vtamR", recursive = TRUE, force = TRUE)
file.exists("C:/Users/emese/AppData/Local/R/win-library/4.4/vtamR") 
unlink("~/R/x86_64-pc-linux-gnu-library/4.0/vtamR", recursive = TRUE, force = TRUE)
file.exists("~/R/x86_64-pc-linux-gnu-library/4.0/vtamR") 
# Clean and regenerate documentation:
devtools::clean_dll()
devtools::document()

devtools::install(build_vignettes = FALSE, force = TRUE, dependencies = TRUE)
```

### Build vignettes
#### It creates Rmd and html in doc
```{r}
setwd("/home/meglecz/vtamR")
setwd("C:\\Users\\emese\\vtamR")
library("devtools")
devtools::build_vignettes()
```

#### Make an html with toc float

copy Rmd to web_doc modify TOC
```{r}
  html_document:
    toc: true
    toc_float: true
# knit and put the html to webpage
# make index to link the different pages
```

## static github website
Run all this from vtamR and not from vtamR/divers !!!

```{r}
install.packages("pkgdown")
```

In R, from package root with develop checked out: 
This creates a _pkgdown.yml file and adds docs/ to .Rbuildignore automatically. Skip this step if _pkgdown.yml already exists.
```{r}
usethis::use_pkgdown()
```
Correct the URL manually to url: https://meglecz.github.io/vtamR/


This renders your vignettes (already working locally), builds the function reference from your roxygen docs, and writes the full site into docs/ at the repo root.
Do this every time the pkgdown site should be updated

```{r}
pkgdown::build_site()
```

```{bash}
git add docs/ .Rbuildignore _pkgdown.yml
git commit -m "Build pkgdown site"
git push origin develop
```


On GitHub: 
go to your repo → Settings → Pages → under 'Build and deployment', 
  set 
    Source to 'Deploy from a branch', 
    Branch to develop, 
    folder to /docs. 
  Save.
  
  
GitHub takes a minute or two to build and deploy. 
Then visit https://meglecz.github.io/vtamR/ to confirm it's live. 
If it 404s at first, give it a few minutes and refresh — first deploy can be slow.

When submitting the paper, set doc to main???


## create data release

### set a personal access token

```{r}
  usethis::edit_r_environ()
```

Edit the file to add or update 
```
# GITHUB_PAT=ghp_xxxxxxxxxxxx
```

### install piggyback, No need for users
```{r}
install.packages("piggyback")
library(piggyback)
```

### zip output files
move /home/meglecz/vtamR/vignettes/vtamR_demo_out/zfzr_plate1/ to 
/home/meglecz/vtamR/short_tutorial_output/ and zip

```{r}
zip::zip("short_tutorial_output.zip", files = "short_tutorial_output/", recurse = TRUE)
```

First time only: create the release
```{r}
piggyback::pb_release_create(repo = "meglecz/vtamR", tag = "short-tutorial-output-v1")
```

### Every time the output format changes: overwrite the asset
```{r}
piggyback::pb_upload(
  file = "short_tutorial_output.zip",
  repo = "meglecz/vtamR",
  tag  = "short-tutorial-output-v1"
)
```

### users do
```{r}
tmp_out <- download_tutorial_output()
```


# Users

## Install

#### from github develop
```{r}
if(!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
pak::pkg_install("meglecz/vtamR@develop", dependencies=TRUE)

if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("meglecz/vtamR@develop", dependencies = TRUE)
```

### Re-install from github develop
```{r}
remotes::install_github(
  repo = "meglecz/vtamR",       # Replace with the repository name
  ref = "develop",       # Replace with the branch name
  upgrade = TRUE             # Optional: to upgrade if already installed
)
```


## read vignettes
```{r}
library("vtamR") 

vignette(package = "vtamR")
vignette("installation") 
vignette("tutorial-vtamr-pipeline") 
vignette("from-fastq-to-df") 
vignette("make-mock-composition-file") 

```


## info package

### Lists all  objects from namespace:
```{r}
getNamespaceExports("vtamR")
l <- getNamespaceExports("vtamR")
write.csv(l, "vtam_functions.csv")
```

### Shows the package help page with an index of functions:
```{r}
help(package = "vtamR")
?vtamR

apropos("filter")
```

