---
title: "Quick Start"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=TRUE,
  cache=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

**Package version:** `r packageVersion("vtamR")`

# Set Paths to Third-Party Programs 
Adjust path according to you installation, or leave this out, if external programs are in your path.

```{r, eval=FALSE}
options(vtamR.vsearch_path = "C:/Users/Public/vsearch-2.23.0-win-x86_64/bin/vsearch")
options(vtamR.blast_path = "C:/Users/Public/blast-2.16.0+/bin/blastn")
options(vtamR.cutadapt_path = "C:/Users/Public/cutadapt")
options(vtamR.swarm_path = "C:/Users/Public/swarm-3.1.5-win-x86_64/bin/swarm")
options(vtamR.pigz_path = "C:/Users/Public/pigz-win32/pigz")
```



```{r}
## Load Library
library(vtamR)
library(dplyr)

## Demo dataset
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package = "vtamR")
mock_ncbi_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package = "vtamR")
asv_list <-  system.file("extdata/demo/asv_list.csv", package = "vtamR")

taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")

### Check the Format of Input Files
check_file_info(file = fastqinfo, dir = fastq_dir, file_type = "fastqinfo")
check_file_info(file = asv_list, file_type = "asv_list")

## Output directory and log file
outdir <- file.path("vtamR_demo_out", "QuickStart_mfzr_plate1")
options(vtamR.log_file = file.path(outdir, "run_info", "vtamR_log.csv"))

# Pre-process reads
## Demultiplex
demultiplexed_dir <- file.path(outdir, "1_demultiplexed")
fastqinfo_df <- demultiplex_and_trim_fastq(
  fastqinfo = fastqinfo,
  fastq_dir = fastq_dir,
  outdir = demultiplexed_dir,
  primer_to_end = FALSE,
  check_reverse = TRUE,
  compress = FALSE
)

## Randomly subsample sequencing reads
randomseq_dir <- file.path(outdir, "2_random_sampled")
fastqinfo_df <- random_sample_batch_by(
  info = fastqinfo_df,
  by = "replicate",
  dir = demultiplexed_dir,
  outdir = randomseq_dir,
  N = 40000,
  compress = FALSE
  ) 

## Merge Fastq Pairs and Quality Filter {#merge}
merged_dir <- file.path(outdir, "3_merged")
sampleinfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo_df,
  fastq_dir = randomseq_dir,
  outdir = merged_dir,
  fastq_maxee = 1,
  fastq_maxns = 0,
  fastq_allowmergestagger = T
)

## Dereplicate
outfile <- file.path(outdir, "4_filter", "1_before_filter.csv")
updated_asv_list <- file.path(outdir, "ASV_list_with_IDs.csv")
read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df, # input df (output of demultiplex_and_trim_fasta)
  dir = merged_dir, # directory with input fasta files
  outfile = outfile, # write the output data frame to this file
  input_asv_list = asv_list
  )

# Filtering
## Initialize stat data frame
stat_df <- get_stat(read_count_df, stat_df = NULL, stage = "input", params = NA)

## Denoising with SWARM {#denoise}
split_clusters <- TRUE
by_sample <- TRUE
outfile <- file.path(outdir, "4_filter", "2_swarm_denoise.csv")
read_count_df <- denoise_by_swarm(
  read_count = read_count_df,
  outfile=outfile,
  by_sample=by_sample,
  split_clusters=split_clusters,
  quiet=TRUE
  )
param <- paste("by_sample:", by_sample, "split_clusters:", split_clusters, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="swarm_denoise", params=param)

## Filter ASVs
### Filter out ASVs with Low Total Read Count (`filter_asv_global`) {#filter_asv_global}
global_read_count_cutoff = 2
outfile <- file.path(outdir, "4_filter", "3_filter_asv_global.csv")
read_count_df <- filter_asv_global(
  read_count = read_count_df, 
  outfile = outfile,
  cutoff = global_read_count_cutoff
  )
param <- paste("global_read_count_cutoff:", global_read_count_cutoff, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_asv_global", params=param)

### Update the asv_list
updated_asv_list <- file.path(outdir, "updated_ASV_list_mfzr.csv")

update_asv_list(
  asv_list1 = read_count_df,
  asv_list2 = asv_list,
  outfile = updated_asv_list)

### Filter ASVs Based on Read Length (`filter_indel`) {#filter_indel}
outfile <- file.path(outdir, "4_filter", "4_filter_indel.csv")
read_count_df <- filter_indel(
  read_count = read_count_df,
  outfile = outfile
  )
stat_df <- get_stat(read_count_df, stat_df, stage="filter_indel", params=NA)

### Filter ASVs with Stop Codons (`filter_stop_codon`) {#filter_stop_codon}
outfile <- file.path(outdir, "4_filter", "5_filter_stop_codon.csv")
genetic_code = 5
read_count_df <- filter_stop_codon(
  read_count = read_count_df, 
  genetic_code = genetic_code,
  outfile = outfile
  )
param <- paste("genetic_code:", genetic_code, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_stop_codon", params=param)

### Filter Potential Contaminant ASVs (`filter_contaminant`) {#filter_contaminant}
outfile <- file.path(outdir, "4_filter", "6_filter_contaminant.csv")
conta_file <- file.path(outdir, "4_filter", "potential_contaminants.csv")
read_count_df <- filter_contaminant(
  read_count = read_count_df, 
  sampleinfo = sampleinfo_df,
  conta_file = conta_file,
  outfile = outfile
  )
stat_df <- get_stat(read_count_df, stat_df, stage="filter_contaminant", params=NA)

### Filter Chimeric ASVs (`filter_chimera`) {#filter_chimera}
outfile <- file.path(outdir, "4_filter", "7_filter_chimera.csv")
abskew=2
by_sample = TRUE
sample_prop = 0.8
read_count_df <- filter_chimera(
  read_count = read_count_df, 
  outfile = outfile,
  by_sample = by_sample, 
  sample_prop = sample_prop, 
  abskew = abskew
  )
param <- paste("by_sample:", by_sample, "sample_prop:", sample_prop, "abskew:", abskew, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_chimera", params=param)

## Filter Aberrant Replicates (`filter_replicate`) {#filter_replicate}
renkonen_within_df <- compute_renkonen_distances(
  read_count = read_count_df, 
  compare_all = FALSE)

### Visualize Renkonen Distances
plot_renkonen_distance_density(
  df = renkonen_within_df)

plot_renkonen_distance_barplot(
  df = renkonen_within_df, 
  sampleinfo = sampleinfo_df, 
  x_axis_label_size = 6
  )

### Determine Cutoff for Replicate Filtering
outfile <- file.path(outdir, "4_filter", "8_filter_replicate.csv")
renkonen_distance_quantile <- 0.9
read_count_df <- filter_replicate(
  read_count = read_count_df, 
  outfile = outfile,
  renkonen_distance_quantile = renkonen_distance_quantile
  )
param <- paste("renkonen_distance_quantile:", renkonen_distance_quantile, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_replicate", params=param)

## Create `mock_composition` File {#match_variants_to_mock_species}
outdir_mock <- file.path(outdir, "mock_composition")
mock_template <- match_variants_to_mock_species(
  read_count = read_count_df,
  fas = mock_ncbi_fasta, # fasta with reference sequences on the mock or closely related species
  taxonomy = taxonomy,
  sampleinfo = sampleinfo_df,
  outdir = outdir_mock
  )
mock_composition <- file.path(outdir_mock, "mock_composition_template_to_check.csv")
mock_composition_df <- read.csv(mock_composition)

## Filter occurrences
### Filter Low-Abundance Occurrences within Samples (`filter_occurrence_sample`) {#filter_occurrence_sample}
outfile = file.path(outdir, "4_filter", "suggest_parameters", "sample_cutoff.csv")
suggest_sample_cutoff_df <- suggest_sample_cutoff(
  read_count = read_count_df,
  mock_composition = mock_composition,
  outfile = outfile
  )

#### Filter Low-Abundance Occurrences within Samples 
sample_cutoff <- 0.004
outfile <- file.path(outdir, "4_filter", "9_filter_occurrence_sample.csv")
read_count_df <- filter_occurrence_sample(
  read_count = read_count_df, 
  cutoff = sample_cutoff, 
  outfile = outfile
  )
param <- paste("cutoff:", sample_cutoff, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_occurrence_sample", params=param)

### Filter Occurrences by Replicate Consistency (filter_min_replicate-1) {#filter_min_replicate-1}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "10_filter_min_replicate.csv")
read_count_df <- filter_min_replicate(
  read_count = read_count_df, 
  cutoff = min_replicate_number, 
  outfile = outfile
  )
param <- paste("cutoff:", min_replicate_number, sep=" ")
stat_df <- get_stat(read_count_df, stat_df, stage="filter_min_replicate", params=param)

### Filter Low-Abundance Occurrences (`filter_occurrence_variant` and `filter_occurrence_read_count`) {#filter_occurrence_variant}
outdir_suggest = file.path(outdir, "4_filter", "suggest_parameters")
suggest_variant_readcount_cutoffs_df <- suggest_variant_readcount_cutoffs(
  read_count = read_count_df,
  mock_composition = mock_composition,
  sampleinfo = sampleinfo_df,
  outdir = outdir_suggest, 
  min_replicate_number = 2
  )

#### Apply `filter_occurrence_variant` and `filter_occurrence_read_count` {#filter_occurrence_read_count}
outfile <- file.path(outdir, "4_filter", "11_filter_occurrence_variant.csv")
variant_cutoff = 0.001
read_count_df_variant <- filter_occurrence_variant(
  read_count = read_count_df, 
  cutoff = variant_cutoff, 
  outfile = outfile
  )
param <- paste("cutoff:", variant_cutoff, sep=" ")         
stat_df <- get_stat(read_count_df_variant, stat_df, stage="filter_occurrence_variant", params=param)

read_count_cutoff <- 10
outfile <- file.path(outdir, "4_filter", "12_filter_occurrence_read_count.csv")
read_count_df_rc <- filter_occurrence_read_count(
  read_count = read_count_df, 
  cutoff = read_count_cutoff, 
  outfile = outfile
  )
param <- paste("cutoff:", read_count_cutoff, sep=" ")        
stat_df <- get_stat(read_count_df_rc, stat_df, stage="filter_occurrence_read_count", params=param)

outfile <- file.path(outdir, "4_filter", "13_pool_filters.csv")
read_count_df <- pool_filters(
  read_count_df_rc, 
  read_count_df_variant,    
  outfile=outfile
  )
stat_df <- get_stat(read_count_df, stat_df, stage="pool_filters", params=NA)
              
# delete temporary data frames
rm(read_count_df_rc)
rm(read_count_df_variant)

### Filter Occurrences by Replicate Consistency (`filter_min_replicate`-2) {#filter_min_replicate-2}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "14_filter_min_replicate.csv")
read_count_df <- filter_min_replicate(
  read_count = read_count_df, 
  cutoff = min_replicate_number, 
  outfile = outfile
  )
param <- paste("cutoff:", min_replicate_number, sep=" ")  
stat_df <- get_stat(read_count_df, stat_df, stage="filter_min_replicate_2", params=param)

## Pool Replicates {#pool_replicates}
outfile <- file.path(outdir, "4_filter", "15_pool_replicates.csv")
read_count_sample_df <- pool_replicates(
  read_count = read_count_df,
  method = "mean",   # Aggregate replicates using the mean read count
  outfile = outfile
)

## Performance Metrics {#performance-metrics}
false_negatives <- file.path(outdir, "4_filter", "false_negatives.csv")
performance_metrics <- file.path(outdir, "4_filter", "performance_metrics.csv")
known_occurrences <- file.path(outdir, "4_filter", "known_occurrences.csv")
results <- classify_control_occurrences(
  read_count = read_count_sample_df, 
  sampleinfo = sampleinfo_df, 
  mock_composition = mock_composition, 
  known_occurrences = known_occurrences, 
  false_negatives = false_negatives,
  performance_metrics = performance_metrics)
# give explicit names to the 3 output data frames
known_occurrences_df <- results[[1]]
false_negatives_df <- results[[2]]
performance_metrics_df <- results[[3]]

## Print Summary Statistics {#print_stats}
write.csv(stat_df, file = file.path(outdir, "run_info", "summary.csv"))

# Taxonomic Assignment
outfile <- file.path(outdir, "4_filter", "assign_taxonomy_ltg.csv")
asv_tax <- assign_taxonomy_ltg(
  asv = read_count_sample_df, 
  taxonomy = taxonomy, 
  blast_db = blast_db, 
  outfile = outfile)

# Clustering (mOTUs) {#cluster_asv}
### Plot Pairwise Identity Distribution
outfile <- file.path(outdir, "5_cluster", "pairwise_identities_vsearch.csv")
plotfile <- file.path(outdir, "5_cluster", "plot_pairwise_identity_vsearch.png")
plot_vsearch <- plot_pairwise_identity_vsearch(
  read_count = read_count_sample_df,
  identity_min = 0.9,
  identity_max = 0.99,
  identity_increment = 0.01,
  min_id = 0.8, 
  outfile = outfile, 
  plotfile = plotfile)
print(plot_vsearch)

### Classify Clusters Based On Taxonomy
outfile <- file.path(outdir, "5_cluster", "classification_vsearch.csv")
plotfile <- file.path(outdir, "5_cluster", "plot_cluster_classification_vsearch.png")
scatterplot_vsearch <- plot_cluster_classification(
  read_count = read_count_sample_df,
  taxa = asv_tax,
  clustering_method = "vsearch",
  cluster_params = c(0.90, 0.91, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99),
  taxlevels = c("species", "genus"),
  outfile = outfile,
  plotfile = plotfile)
print(scatterplot_vsearch)

## Cluster ASVs into mOTUs
outfile <- file.path(outdir, "5_cluster", "16_mOTU_vsearch.csv")
identity <- 0.97 
read_count_samples_df_ClusterSize <- cluster_asv(
  read_count = read_count_sample_df,
  method = "vsearch",
  identity = 0.97,
  group = TRUE,
  by_sample = FALSE,
  outfile = outfile)
stat_df <- get_stat(read_count_samples_df_ClusterSize, 
                   stat_df, 
                   stage="cluster_asv_vsearch",
                   param=identity)

# Output 

## ASV table {#write_asv_table}
outfile=file.path(outdir, "motu_table_taxa.csv")
asv_table_df <- write_asv_table(
  read_count = read_count_sample_df, 
  outfile = outfile, 
  asv_tax = asv_tax, 
  sampleinfo = sampleinfo_df, 
  add_empty_samples = FALSE, 
  add_sums_by_sample = TRUE, 
  add_sums_by_asv = TRUE, 
  add_expected_asv = TRUE, 
  mock_composition = mock_composition)

## Convert output to a `phyloseq` object{#phyloseq}
# Create a data frame containing sample metadata
samples <- sampleinfo_df %>%
  select(sample, sample_type, habitat) %>%
  distinct()
phy_object <- format_for_phyloseq(
  otu = read_count_sample_df,
  tax = asv_tax,
  samples = samples,
  outfile = file.path(outdir, "phyloseq", "phyloseq.rds")
)

## Convert output for the `vegan` package{#vegan}
vegan_dfs <- format_for_vegan(
  otu = read_count_sample_df,
  tax = asv_tax,
  sample_type = sampleinfo_df,
  rm_coltrol = TRUE,
  outfile_motu = file.path(outdir, "vegan", "motu_table_vegan.csv"), 
  outfile_taxa = file.path(outdir, "vegan", "tax_table_vegan.csv")
)
vegan_motu_df <- vegan_dfs[[1]]
vegan_tax_df <- vegan_dfs[[2]]

# Document your workflow
## Record R and package versions{#package-versions}
package_versions <- file.path(outdir, "run_info", "R_package_versions.csv")
collect_package_info(package_versions)

## MIEM
miem_dir <- file.path(outdir, "run_info", "MIEM")
miem_bioinformatics(
  r_versions = package_versions,
  outfile =  file.path(miem_dir, "MIEM_bioinfo.csv")
  )

## MIEM Sequencing Summary
info_files = c(
  fastqinfo,
  file.path(outdir, "1_demultiplexed/fastqinfo.csv"),
  file.path(outdir, "2_random_sampled/fastqinfo.csv"),
  file.path(outdir, "3_merged/fastainfo.csv")
)
read_count_files = c(
  file.path(outdir, "4_filter/1_before_filter.csv"),
  file.path(outdir, "4_filter/2_swarm_denoise.csv"),
  file.path(outdir, "4_filter/14_filter_min_replicate.csv"),
  file.path(outdir, "4_filter/15_pool_replicates.csv"),
  file.path(outdir, "5_cluster/16_mOTU_vsearch.csv")
)
taxa = file.path(outdir, "4_filter/assign_taxonomy_ltg.csv")

miem_results(
  info_files = info_files, 
  fastq_dir = fastq_dir, 
  read_count_files = read_count_files, 
  taxa = taxa, 
  outdir = miem_dir, 
  mock_composition = mock_composition
  )
```