---
title: "Oool Multiple Datasets"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{Pool Multiple Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=TRUE,
  cache=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

**Package version:** `r packageVersion("vtamR")`


The [previous tutorial]() walked you through a detailed analysis of a single dataset, 
typically consisting of sequences from one sequencing plate.

In this tutorial, I will show you how to combine **datasets targeting the same marker**, 
as well as results from the **same samples obtained using multiple strongly overlapping markers**.

# One Marker

The [main tutorial]() presents the pipeline for analysing the first demo dataset, **mfzr_plate1**. 
The second plate for the MFZR marker, **mfzr_plate2**, is covered in the [short tutorial]().

The two pipelines differ in the order of the preprocessing steps and demonstrate 
different approaches to the analysis. However, the filtering steps are identical in both pipelines.

Here, we continue from the outputs of these two tutorials and show how to pool 
datasets generated from the same marker.

When pooling data from multiple plates, we recommend using the 
**filtered ASV tables** (e.g., `15_pool_replicates_xxx.csv`) 
**before clustering** the sequences into mOTUs. This ensures that clustering 
is performed consistently across datasets and avoids differences in cluster 
centroids between individual runs.

These files are provided in a [long format](), with the columns
`asv_id`, `sample`, `replicate` (optional), `read_count`, and `asv`. 
This format makes them particularly easy to combine.

In this example, we will:

* Pool the filtered ASVs from the two MFZR plates.
* Cluster the pooled ASVs into mOTUs.
* Assign taxonomy to the resulting mOTU centroids.

Load library
```{r, eval=TRUE}
library(vtamR)
```

Define the **output directory and log file**:

```{r, eval=TRUE}
library(vtamR)
outdir <- file.path("vtamR_demo_out", "pooled_mfzr_plates") 
options(vtamR.log_file = file.path(outdir, "vtamR_log.csv"))
```

Define the **taxassign database**
```{r, eval=TRUE}
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")
```

## Pool Datasets

Let's first specify the filtered ASV tables from the two plates:

```{r, eval=TRUE}
pooled_replicates_mfzr1 <- system.file(
  "extdata/demo/15_pool_replicates_mfzr_plate1.csv", 
  package = "vtamR"
)

pooled_replicates_mfzr2 <- system.file(
  "extdata/demo/15_pool_replicates_mfzr_plate2.csv", 
  package = "vtamR"
)

filtered_files <- c(
  pooled_replicates_mfzr1, 
  pooled_replicates_mfzr2
)
```

We can then pool the two datasets:

```{r, eval=TRUE}
outfile <- file.path(outdir, "filtered_asv.csv")

pooled_read_count_df <- pool_datasets(
  files = filtered_files, 
  outfile = outfile, 
  method = "mean"  # If a sample appears on multiple plates, use the mean read count
)
```

## Make mOTUs from the Pooled Data

Next, cluster the pooled ASVs into mOTUs:

```{r, eval=TRUE}
outfile <- file.path(outdir, "motu.csv")
identity <- 0.97 

read_count_motu <- cluster_asv(
  read_count = pooled_read_count_df,
  method = "vsearch",
  identity = identity,
  group = TRUE,
  outfile = outfile
)
```

## Assign Taxa to the Pooled Data

Although we could reuse the taxonomic assignments from the individual plates, 
it is usually simpler to rerun the taxonomic assignment on the pooled dataset. 
This ensures that the assignment is performed on the final set of pooled mOTU centroids.

```{r, eval=TRUE}
outfile <- file.path(outdir, "tax_assign_ltg.csv")

asv_tax_pooled <- assign_taxonomy_ltg(
  asv = read_count_motu, 
  taxonomy = taxonomy, 
  blast_db = blast_db, 
  outfile = outfile
)
```

## Print the ASV Table

Finally, generate a table containing the mOTU read counts and their taxonomic assignments:

```{r, eval=TRUE}
outfile <- file.path(outdir, "motu_table_taxa.csv")

asv_table_df <- write_asv_table(
  read_count = read_count_motu, 
  outfile = outfile, 
  asv_tax = asv_tax_pooled, 
  add_sums_by_sample = TRUE, 
  add_sums_by_asv = TRUE
)
```


# More than One Overlapping Marker

Using two or more **overlapping markers** on the same samples can 
improve taxonomic coverage by targeting complementary regions of the marker.

Each marker should first be analyzed separately. The results can then be 
combined by aggregating ASVs that are **identical within their overlapping region**.

The first plate of the ZFZR marker can be processed in the same way as the first MFZR plate.

Here, we use the output from the ZFZR marker analysis, provided as a demo file. 
The corresponding results are included in the package:

```{r, eval=TRUE}
mfzr_filtered <- system.file(
  "extdata/demo/15_pool_replicates_mfzr_plate1.csv", 
  package = "vtamR"
)

zfzr_filtered <- system.file(
  "extdata/demo/15_pool_replicates_zfzr_plate1.csv", 
  package = "vtamR"
)
```

Let's define a new output directory and log file:

```{r, eval=TRUE}
outdir <- file.path("vtamR_demo_out", "pooled_markers") 
options(vtamR.log_file = file.path(outdir, "vtamR_log.csv"))
```

## Pool Markers

When analyzing the different datasets, we consistently used the 
`ASV_list_with_IDs.csv` generated in the previous runs. 
As a result, the ASV IDs are compatible across datasets and can be used directly.

If the ASV IDs are **not compatible** — for example, if the same `asv_id` 
refers to different sequences in different markers—you can resolve this by 
assigning a numerical identifier to each marker. 
In this example, we use `1` for MFZR and `2` for ZFZR.

These marker identifiers are appended to the ASV IDs, 
ensuring that they remain unique and preventing confusion between markers.

```{r, eval=TRUE}
input_files <- c(mfzr_filtered, zfzr_filtered)
marker_ids <- c(1, 2)

outfile <- file.path(outdir, "filtered_asv.csv") 
asv_with_centroids <- file.path(outdir, "filtered_asv_with_centroids.csv")

pooled_markers <- pool_markers(
  files = input_files,
  marker_ids = marker_ids,
  outfile = outfile,  # ASVs identical in their overlapping region are merged into a single line
  asv_with_centroids = asv_with_centroids,  # same as input, with added centroid_id and centroid
  method = "mean"  # mean read count across pooled ASVs
)
```

## Make mOTU  - Taxassign - mOTU table

As with multiple datasets for a single marker, we will cluster the ASVs 
from the pooled dataset into mOTUs, then assign the sequences to taxa, 
and finally create an mOTU table.

```{r, eval=TRUE}
#### Cluster
outfile <- file.path(outdir, "motu.csv") 
identity <- 0.97 
read_count_motu <- cluster_asv(
  read_count = pooled_markers,
  method = "vsearch",
  identity = 0.97,
  group = TRUE,
  outfile = outfile
  )

#### TaxAssign
outfile <- file.path(outdir, "assign_taxonomy_ltg.csv") 

asv_tax_pooled <- assign_taxonomy_ltg(
  asv = read_count_motu, 
  taxonomy = taxonomy, 
  blast_db = blast_db, 
  outfile = outfile
  )

#### Print mOTU table
outfile <- file.path(outdir, "motu_table_taxa.csv") 

asv_table_df <- write_asv_table(
  read_count = read_count_motu, 
  outfile = outfile, 
  asv_tax = asv_tax_pooled, 
  add_sums_by_sample = T, 
  add_sums_by_asv = T
  )
```


