---
title: "Troubleshooting"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=FALSE,
  cache=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

**Package version:** `r packageVersion("vtamR")`

# Troubleshooting

## Pool different runs of the same samples before filtering

Pooling different sequencing runs is useful when the same samples were sequenced 
in multiple runs, for example if one sequencing run was of lower quality or if 
technical replication was performed. Pooling can be done before filtering
to create a unified dataset.

### Option 1

The simplest approach is to concatenate the corresponding input files (FASTQ or FASTA).

The `concatenate_files` function reads files from multiple directories 
and concatenates the contents of files with the same filename into a single output file.

**Important**: Before using this function, make sure that the sequence IDs in 
the files to be concatenated are unique. Duplicate IDs can cause issues in downstream analyses.

```{r, eval=FALSE}
concatenate_files(
   dirs = c("run1", "run2", "run3"),
   outdir = "pooled",
   pattern = ".fastq$"
)
```

### Option 2

Safer but a bit more complicated option.

#### 1. Process each run separately up to dereplication

For each sequencing run, perform the following steps:

* Merge fastq read pairs to assemble paired-end reads
* Demultiplex the reads
* Trim primers and adapters

Follow the steps in [from-fastq-to-df](from-fastq-to-df.html), adapting to 
your wet lab and sequencing protocol. Store the trimmed, merged, and demultiplexed 
FASTA files in a separate directory for each run.

#### 2. Dereplicate each run independently, while keeping a shared ASV list

Dereplicate each run separately, but ensure that ASV identifiers (`asv_id`) 
remain consistent across runs by:

* Using the `output_asv_list` from the first `dereplicate()` call
* Providing it as `input_asv_list` to the next run
* Passing the updated list to subsequent runs

This approach guarantees that previously detected ASVs retain the same `asv_id` 
and newly detected ASVs receive new identifiers. 
Harmonizing ASV IDs across datasets is essential.

```{r, eval=FALSE}
# Run 1
updated_asv_list1 <- file.path(outdir, "ASV_list_with_IDs_1.csv")

read_count_df1 <- dereplicate(
  sampleinfo = sampleinfo_df1,
  dir = demultiplexed_dir1,
  input_asv_list = "",  # or an existing ASV list
  output_asv_list = updated_asv_list1
)

# Run 2
updated_asv_list2 <- file.path(outdir, "ASV_list_with_IDs_2.csv")

read_count_df2 <- dereplicate(
  sampleinfo = sampleinfo_df2,
  dir = demultiplexed_dir2,
  input_asv_list = updated_asv_list1,
  output_asv_list = updated_asv_list2
)

# Run 3
updated_asv_list3 <- file.path(outdir, "ASV_list_with_IDs_3.csv")

read_count_df3 <- dereplicate(
  sampleinfo = sampleinfo_df3,
  dir = demultiplexed_dir3,
  input_asv_list = updated_asv_list2,
  output_asv_list = updated_asv_list3
)
```

#### 3. Pool the dereplicated datasets

After dereplication and harmonization of ASV identifiers, the resulting 
`read_count_df` objects can be combined.

```{r, eval=FALSE}
read_count_df <- rbind(read_count_df1, read_count_df2, read_count_df3) %>%
  group_by(asv_id, sample, replicate, asv) %>%
  summarize(read_count = sum(read_count), .groups = "drop")
```

This pooling step:

* Combines multiple runs
* Sums read counts for identical ASV–sample–replicate combinations
* Produces a unified dataset ready for downstream filtering

## `filter_occurrence_variant` can eliminate most occurrences of a frequent ASV

The `filter_occurrence_variant` function removes all occurrences where 
the proportion of reads of an ASV in a sample-replicate relative to its total 
read count is below the specified cutoff:

`(number of reads of the ASV in a sample-replicate) / (total number of read of the ASV) < cutoff`

When 

- an ASV occurs in many samples and 
- the dataset contains numerous samples and 
- the cutoff is relatively high

most occurrences of that ASV may fall below the cutoff, causing its total read 
count to drop sharply during this filtering step.

If the proportion of reads retained for an ASV after filtering is less than 
`min_read_count_prop`, vtamR issues a warning. This should be carefully inspected:

* The ASV might represent a general contaminant, in which case it may be 
appropriate to remove it entirely from the dataset.
* Otherwise, consider lowering the cutoff value in `filter_occurrence_variant`
to retain more occurrences of the ASV.

## Memory issues for running SWARM on the whole data set

When working with very large datasets, running SWARM on the entire dataset at 
once can cause memory issues. 

Possible workarounds:

* Run SWARM sample by sample (`by_sample = TRUE`).
* Run `global_read_count` to remove singletons before SWARM.

Both may slightly reduce clustering efficiency but allow the analysis to complete.

## Memory issues in `merge_fastq_pairs` on Windows

If very large input fastq files fail to merge, try using **uncompressed input files**.

## No or few sequences passing `merge_fastq_pairs`

If your amplicon is shorter than the reads, enable staggered merging by setting 
`fastq_allowmergestagger = TRUE`.

Certainly! Here’s a clearer, more tutorial-friendly rewrite of your section:


## No or Few Sequences Passing `demultiplex_and_trim_fasta`

By default, the `demultiplex_and_trim_fasta` function searches for tags and primers 
only at the very ends of the reads (`tag_to_end = TRUE`, `primer_to_end = TRUE`). 
This setting is both efficient and appropriate for most use cases.

However, if you used sequencing spacers, the primers or tags might be located 
a few nucleotides away from the ends of your sequences. In such cases, 
you need to adjust these settings to ensure the function searches for 
tags and primers in the correct positions.

For more details, refer to the function documentation: `?demultiplex_and_trim_fasta`.


