---
title: "From fastq to data frame"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{From fastq to data frame }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


According to your wet lab and sequencing protocol, each FASTQ file may contain one or more [sample-replicates](tutorial-vtamr-pipeline.html#glossary). In addition, sequences may include [tags](tutorial-vtamr-pipeline.html#glossary) 
(used for [demultiplexing](tutorial-vtamr-pipeline.html#glossary)) 
as well as primer sequences.

In the following sections, we present **three common scenarios** for generating the [read_count_df](tutorial-vtamr-pipeline.html#read_count_df) data frame, 
which serves as the input for the filtering steps.

# From fastq to data frame

**Package version:** `r packageVersion("vtamR")`

## Set up

*Load library*
```{r setup}
library(vtamR)
```


If the third-party tools are not in your PATH, set the corresponding package options and specify the correct path to each tool.

**Adapt the path** to third party programs according to your installation
(See [Installation](installation.html)).

```{r, eval=FALSE}
options(vtamR.vsearch_path = "C:/Users/Public/vsearch-2.23.0-win-x86_64/bin/vsearch")
options(vtamR.cutadapt_path = "C:/Users/Public/cutadapt")
options(vtamR.pigz_path = "C:/Users/Public/pigz-win32/pigz")
```

```{r, echo = FALSE, eval=TRUE}
options(vtamR.vsearch_path = "~/miniconda3/envs/vtam/bin/vsearch")
options(vtamR.cutadapt_path = "~/miniconda3/envs/vtam_benchmark/bin/cutadapt")
options(vtamR.pigz_path = "~/miniconda3/envs/vtam/bin/pigz")
```

See the [detailed tutorial on how to set the PATH](tutorial-vtamr-pipeline.html#set-path) for more information.

`compress = FALSE` In the tutorial we will use uncompressed files, but see
[To compress or not to compress](tutorial-vtamr-pipeline.html#to-compress-or-not-to-compress)
for the best strategy for you. 
See also [Files to keep](tutorial-vtamr-pipeline.html#files-to-keep) on how to handle the 
large intermediate files that should be compressed or deleted once the analyses are complete.


## Case 1 - One sample per fastq - No tag - No primer

In this scenario, each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads have been trimmed from all artificial add-ons, such as
adapters, tags, indices and also from primers.

Read pairs should be **quality filtered**, **merged** and written to 
**fasta format**.
This can be done by the `merge_fastq_pairs` function. 

See the help (`?merge_fastq_pairs`) for setting the 
correct parameters for quality filtering.

### Set Input

Set output directory and log life
```{r}
outdir <- "vtamR_demo_case1"
# set package option to define the log_file
options(vtamR.log_file =  file.path(outdir, "vtamR_log.csv"))
```

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input1}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo1.csv", package = "vtamR")
```

### Merge Fastq File Pairs and Quality Filter Reads

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): is either a csv file, or a data frame. The key 
information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged. 
The `tag_fw`, `primer_fw`, `tag_rv`, `primer_rv` are irrelevant in this case, 
just fill them with `NA`.
* [sampleinfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `merge_fastq_pairs`. It is the updated 
version of fastqinfo, where fastq file names have been replaced by fasta file 
names and the read counts are included for each file. 

```{r merge1}
merged_dir <- file.path(outdir, "merged")

sampleinfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo, 
  fastq_dir = fastq_dir, # Directory containing the input FASTQ files.
  outdir = merged_dir, # Name of the output directory
  compress = FALSE
  )
```


### Dereplicate reads

The fasta files produced by `merge_fastq_pairs` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

* [read_count_df]((tutorial-vtamr-pipeline.html#read_count_df): Data frame with
`asv`, `asv_id`,`sample`, `replicate`, (optional) and `read_count` columns.
It is also written to CSV file if the `outfile` is defined.

```{r derelicate1}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df, 
  dir = merged_dir, # Directory containing the input FASTA files.
  outfile = outfile
  )
```


## Case 2 - One Sample per Fastq - Primer - No Tag

This is one of the most frequent case. 
Each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads has been trimmed from all artificial add-ons, such as
adapters, tags, BUT they still have the primers.

Read pairs should be **quality filtered**, **merged** and written to **fasta format** by 
`merge_fastq_pairs` function as in the previous section.

Then the `trim_primer` function will **trim the primers** from the reads. 
See the help (`?trim_primer`) for setting the correct parameters for primer trimming.

### Set Input

**Set output directory and log life**
```{r}
outdir <- "vtamR_demo_case2"
# set package option to define the log_file
options(vtamR.log_file =  file.path(outdir, "vtamR_log.csv"))
```

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input2}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo2.csv", package = "vtamR")
```

### Merge Fastq File Pairs and Quality Filter Reads

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame.  
The key information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged. 
The `primer_fw`, `primer_rv` columns are irrelevant in this case, just fill them with `NA`.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): is the output of `merge_fastq_pairs`. 
It is the updated version of `fastqinfo`, where fastq file names have been 
replaced by fasta file names containing the merged sequences.

```{r merge2}
merged_dir <- file.path(outdir, "merged")

fastainfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo, 
  fastq_dir = fastq_dir, # directory of input FASTQ files
  outdir = merged_dir, # Name of the output directory
  compress = FALSE
  )
```

### Trim Primers

* `sampleinfo_df`: is  updated version of fastainfo_df. 
This data frame and the files listed in it are the input for `dereplicate`.

```{r primer-trim}

demultiplexed_dir <- file.path(outdir, "demultiplexed")
sampleinfo_df <- trim_primers(
  fastainfo = fastainfo_df, 
  fasta_dir = merged_dir, # Directory containing the input FASTA files (created by `merge_fastq_pairs`).
  outdir = demultiplexed_dir,  # Name of the output directory
  check_reverse = TRUE, # Check the reverse complementary strand as well
  primer_to_end = FALSE, 
  compress = FALSE
  )
```

### Dereplicate

The fasta files produced by `trim_primers` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

* [read_count_df]((tutorial-vtamr-pipeline.html#read_count_df): Data frame with
`asv`, `asv_id`,`sample`, `replicate`, (optional) and `read_count` columns.
It is also written to CSV file if the `outfile` is defined.

```{r derelicate2}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df, 
  dir = demultiplexed_dir, 
  outfile = outfile
  )
```

## Case 3 - Several Samples per Fastq - Tags - Primers

In this case, one pair of fastq files contains reads from multiples samples or
sample-replicates, so it is necessary to demultiplex them, and trim from 
tags and primers.

### Case 3a Merge - demultiplex

Read pairs should be **quality filtered**, **merged** and written to **fasta format** as 
in the previous sections.

Then the `demultiplex_and_trim_fasta` function will 
**[demultiplex](tutorial-vtamr-pipeline.html#glossary)** the fasta 
files according to the **[tag](tutorial-vtamr-pipeline.html#glossary)** combinations and **[trim](tutorial-vtamr-pipeline.html#glossary)** 
the primers from the reads. 

See the help (`?demultiplex_and_trim_fasta`) for setting the 
correct parameters for demultiplexing and primer trimming:

#### Set Input

**Set output directory and log life**
```{r}
outdir <- "vtamR_demo_case3a"
# set package option to define the log_file
options(vtamR.log_file =  file.path(outdir, "vtamR_log.csv"))
```

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input3}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package = "vtamR")
```


#### Merge Fastq File Pairs and Quality Filter Reads

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame. 
The key information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `merge_fastq_pairs`. 
It is the updated version of `fastqinfo`, where fastq file names have been 
replaced by fasta file names.

```{r merge3}
merged_dir <- file.path(outdir, "merged")

fastainfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo, 
  fastq_dir = fastq_dir, # Directory containing the input FASTQ files.
  outdir = merged_dir, # Name of the output directory. 
  compress = FALSE
  )
```

#### Demultiplex, Trim off Tags and Primers

* `sampleinfo_df`: Updated version of `fastainfo`. 
This data frame and the files listed in it are the input of the `dereplicate`.

```{r demultiplex3}
demultiplexed_dir <- file.path(outdir, "demultiplexed")

sampleinfo_df <- demultiplex_and_trim_fasta(
  fastainfo = fastainfo_df, 
  fasta_dir = merged_dir, # Directory containing the input FASTA files.
  outdir = demultiplexed_dir, # Name of the output directory. 
  check_reverse = TRUE, # Checks the reverse complementary stand as well
  compress = FALSE
  )
```

#### Dereplicate

The fasta files produced by `demultiplex_and_trim_fasta` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

* [read_count_df]((tutorial-vtamr-pipeline.html#read_count_df): Data frame with
`asv`, `asv_id`,`sample`, `replicate`, (optional) and `read_count` columns.
It is also written to CSV file if the `outfile` is defined.

```{r depelicate3}
outfile <- file.path(outdir, "1_before_filter.csv")
read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df, 
  dir = demultiplexed_dir, # Directory containing the input FASTA files.
  outfile = outfile
  )
```


Here's a polished version with corrected grammar, clearer wording, and improved consistency:

### Case 3b: Demultiplex First, Then Merge

When submitting FASTQ files to the NCBI Sequence Read Archive (SRA), each file 
must contain reads from a single sample-replicate combination and must be trimmed 
of tags/adapters and primers. Therefore, instead of merging reads first and then 
demultiplexing them, as described in the previous section, the order of these 
steps can be reversed.

FASTQ file pairs can be demultiplexed and trimmed of tags and primers using 
the `demultiplex_and_trim_fastq` function. This procedure produces FASTQ file pairs 
that can be submitted directly to the SRA.

The demultiplexed FASTQ files should then be merged and converted to FASTA format 
before proceeding with the filtering steps.

See the help page (`?demultiplex_and_trim_fastq`) for details on setting the 
appropriate parameters for demultiplexing and primer trimming.

#### Input Parameters

**Set output directory and log life**
```{r}
outdir <- "vtamR_demo_case3b"
# set package option to define the log_file
options(vtamR.log_file =  file.path(outdir, "vtamR_log.csv"))
```

The example files used below are included with the `vtamR` package, 
which is why we use `system.file()` to access them in this tutorial.
When using your own data, simply provide the appropriate file and directory 
paths (e.g., `~/vtamR/fastq`). 
Make sure that there are **no spaces in file or directory names**.

```{r set-input4}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package = "vtamR")
```

#### Demultiplex, Trim off Tags and Primers and keep fastq format

* **`fastqinfo`** ([fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo)): 
Either a CSV file or a data frame. For `demultiplex_and_trim_fastq`, the key 
information includes the list of FASTQ file pairs to be demultiplexed,
as well as the primer and tag sequences associated with each sample-replicate combination..

```{r demultiplex4}
demultiplexed_dir <- file.path(outdir, "demultiplexed")

fastqinfo_delultiplexed <- demultiplex_and_trim_fastq(
  fastqinfo = fastqinfo, 
  fastq_dir = fastq_dir, # Directory containing the input FASTQ files.
  outdir = demultiplexed_dir, # Name of the output directory.
  primer_to_end = FALSE,
  check_reverse = TRUE, # Check the reverse-complement strand.
  compress = FALSE,
  quiet = TRUE
  )
```

#### Merge Fastq File Pairs and Quality Filter Reads

* **`sampleinfo_df`**: An updated version of `fastqinfo`. This data frame, 
together with the files listed in it, serves as input for the `dereplicate` step.

```{r merge4}
merged_dir <- file.path(outdir, "merged")

sampleinfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo_delultiplexed, 
  fastq_dir = demultiplexed_dir, # Directory containing the input FASTQ files
  outdir = merged_dir, # Name of the output directory.
  compress = FALSE
  )
```


#### Dereplicate

The fasta files produced by `merge_fastq_pairs` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

* [read_count_df]((tutorial-vtamr-pipeline.html#read_count_df): Data frame with
`asv`, `asv_id`,`sample`, `replicate`, (optional) and `read_count` columns.
It is also written to CSV file if the `outfile` is defined.

```{r depelicate4}
outfile <- file.path(outdir, "1_before_filter.csv")
read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df, 
  dir = merged_dir, # Directory containing the input FASTA files
  outfile = outfile
  )
```

## Subsampling

Subsampling FASTA or FASTQ file pairs is useful when sequences come 
from different sequencing libraries with strongly different numbers of reads. 

For example, in our sample dataset each sequencing library had one replicate 
of all samples. Subsampling each library to a similar number of reads helps 
avoid filtering out occurrences from a replicate that happens to have relatively few reads.

Standardizing read numbers is also a good idea when working with different 
sequencing runs that have strongly different coverage.

The `random_sample_batch` function subsamples (without replacement) reads in 
FASTA files or FASTQ file pairs. You can set any one of the following three parameters:

* `n`: the number of sequences selected from each file
* `p`: the proportion of sequences to be sampled from each file
* `N`: the total number of sequences to be sampled from all input files, 
while keeping the relative read counts among files

This step can be run on the raw input FASTQ file pairs, 
on demultiplexed FASTA or FASTQ file pairs, or on merged FASTA files.

However, if your workflow includes `filter_occurrence_variant` or 
`filter_occurrence_read_count` — which rely on total read counts in negative controls — 
standardizing read numbers across samples is **not recommended**. 
That said, it's still possible to subsample demultiplexed files using the 
proportion of reads to be sampled (`p`) or the total number of reads across 
all samples while keeping relative read counts across samples (`N`).


