---
title: "From fastq to data frame - v0.3.1"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{From fastq to data frame - v0.3.1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


According to your wetlab and sequencing protocol each fastq files can 
contain one or more [sample-replicates](tutorial-vtamr-pipeline.html#glossary), and sequences may or may 
not contain [tags](tutorial-vtamr-pipeline.html#glossary) (for [demultiplexing](tutorial-vtamr-pipeline.html#glossary)) and primer sequences. 
In the following sections I show **3 different scenarios** to obtain the 
[read_count_df](tutorial-vtamr-pipeline.html#read_count_df) data frame, which the input to the filtering steps.

# Set up

*Load library*
```{r setup}
library(vtamR)
```

*Set path to third party programs*
```{r set_path_win, eval=FALSE}
# Example for Windows
cutadapt_path <- "C:/Users/Public/cutadapt"
vsearch_path <- "C:/Users/Public/vsearch-2.23.0-win-x86_64/bin/vsearch"
pigz_path <- "C:/Users/Public/pigz-win32/pigz" # optional to speed file compression
```

```{r set_path_linux}
#  Example for Linux
cutadapt_path <- "~/miniconda3/envs/vtam/bin/cutadapt" # v3.4
vsearch_path <- "~/miniconda3/envs/vtam/bin/vsearch" # v2.15.1
pigz_path <- "~/miniconda3/envs/vtam/bin/pigz" # optional to speed file compression
```

**Adapt the path** to third party programs according to your installation
(See [Installation](installation.html)).

If third party programs are in your `PATH` 
(See [Installation](intsallation.html#install-third-party-programs)), simply omit
the `cutadapt_path`, `pigz_path`, `vsearch_path` argument when calling the `vtamR` functions.

`compress = FALSE` In the tutorial we will use uncompressed files, but see
[To compress or not to compress](tutorial-vtamr-pipeline.html#to-compress-or-not-to-compress) for the best
strategy for you. See also [Files to keep](tutorial-vtamr-pipeline.html#files-to-keep) on how to handle the 
large intermediate files that should be compressed or deleted once the analyses are complete.

# Case 1 - One sample per fastq - no tag - no primer

In this scenario, each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads has been trimmed from all artificial add-ons, such as
adapters, tags, indices and also from primers.

Read pairs should be **quality filtered**, **merged** and written to 
**fasta format**.
This can be done by the `Merge` function. 

See the help (`?Merge`) for setting the correct parameters for quality filtering.

**Set input**

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): is either a csv file, or a data frame. The key 
information for `Merge` is the list of the fastq file pairs that should be merged. 
The `tag_fw`, `primer_fw`, `tag_rv`, `primer_rv` are irrelevant in this case, 
just fill them with `NA`.
* `fastq_dir`: Directory containing the input fastq files.
* [sampleinfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `Merge`. It is the updated 
version of fastqinfo, where fastq file names have been replaced by fasta file 
names and the read counts are included for each file. 
* `outdir`: Name of the output directory.


The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input1}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo1.csv", package = "vtamR")

outdir <- "vtamR_demo_case1"
merged_dir <- file.path(outdir, "merged")
```

**Merge fastq file pairs and quality filter reads**

```{r merge1}
sampleinfo_df <- Merge(fastqinfo, 
                       fastq_dir=fastq_dir, 
                       vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
                       outdir=merged_dir,
                       compress=FALSE)
```


**Dereplicate**

The fasta files produced by `Merge` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?Dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r derelicate1}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- Dereplicate(sampleinfo_df, 
                             dir=merged_dir, 
                             outfile=outfile)
```


# Case 2 - One sample per fastq - primer - no tag

This is one of the most frequent case. 
Each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads has been trimmed from all artificial add-ons, such as
adapters, tags, BUT they still have the primers.

Read pairs should be **quality filtered**, **merged** and written to **fasta format** by 
`Merge` function as in the previous section.

Then the `TrimPrimer` function will **trim the primers** from the reads. 
See the help (`?TrimPrimer`) for setting the correct parameters for primer trimming.

**Set input**

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame.  
The key information for `Merge` is the list of the fastq file pairs that should be merged. 
The `primer_fw`, `primer_rv` columns are irrelevant in this case, just fill them with `NA`.
* `fastq_dir`: Directory containing the input fastq files.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): is the output of `Merge`. 
It is the updated version of `fastqinfo`, where fastq file names have been 
replaced by fasta file names.
* `fasta_dir`: Directory containing the input fasta files for `TrimPrimer`. 
This directory is created by `Merge`.
* If `check_reverse` is TRUE, `TrimPrimer` checks the reverse complementary strand as well.
* `sampleinfo_df`: is  updated version of fastainfo. 
This data frame and the files listed in it are the input for `Dereplicate`.
* `outdir`: Name of the output directory. 

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input2}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo2.csv", package = "vtamR")

outdir <- "vtamR_demo_case2"
merged_dir <- file.path(outdir, "merged")
```

**Merge fastq file pairs and quality filter reads**

```{r merge2}
fastainfo_df <- Merge(fastqinfo, 
                      fastq_dir=fastq_dir, 
                      vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
                      outdir=merged_dir,
                      compress=FALSE)
```

**Trim primers**

```{r primer-trim}

demultiplexed_dir <- file.path(outdir, "demultiplexed")
sampleinfo_df <- TrimPrimer(fastainfo_df, 
                            fasta_dir=merged_dir, 
                            outdir=demultiplexed_dir, 
                            cutadapt_path=cutadapt_path, # can be omitted if CUTADAPT is in the PATH
                            vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
                            check_reverse=T,
                            primer_to_end=F,
                            compress=FALSE)
```

**Dereplicate**

The fasta files produced by `TrimPrimer` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?Dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r derelicate2}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- Dereplicate(sampleinfo_df, 
                             dir=demultiplexed_dir, 
                             outfile=outfile)
```

# Case 3 - Several samples per fastq - tags - primers

In this case, one pair of fastq files contains reads from multiples samples or
sample-replicates, so it is necessary to demultiplex them, and trim from 
tags and primers.

Read pairs should be **quality filtered**, **merged** and written to **fasta format** as 
in the previous sections.

Then the `SortReads` function will **[demultiplex](tutorial-vtamr-pipeline.html#glossary)** the fasta 
files according to the [tag](tutorial-vtamr-pipeline.html#glossary) combinations and **[trim](tutorial-vtamr-pipeline.html#glossary)** 
the primers from the reads. 

See the help (`?SortReads`) for setting the 
correct parameters for demultiplexing and primer trimming:

**Set input**

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame. 
The key information for `Merge` is the list of the fastq file pairs that should be merged.
* `fastq_dir`: Directory containing the input fastq files.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `Merge`. 
It is the updated version of fastqinfo, where fastq file names have been 
replaced by fasta file names.
* `fasta_dir`: Directory containing the input fasta files for `SortReads`. 
This directory is created by `Merge`.
* If `check_reverse` is TRUE, `SortReads` checks the reverse complementary stand as well.
* `sampleinfo_df`: Updated version of fastainfo. 
This data frame and the files listed in it are the input of the `Dereplicate`.
* `outdir`: Name of the output directory. 

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input3}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo.csv", package = "vtamR")

outdir <- "vtamR_demo_case3"
merged_dir <- file.path(outdir, "merged")
demultiplexed_dir <- file.path(outdir, "demultiplexed")
```

**Merge fastq file pairs and quality filter reads**

```{r merge3}
fastainfo_df <- Merge(fastqinfo, 
                      fastq_dir=fastq_dir, 
                      vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
                      outdir=merged_dir,
                      compress=FALSE)
```

**Demultiplex, trim off tags and pimers**
```{r demultiplex}
sampleinfo_df <- SortReads(fastainfo_df, 
                           fasta_dir=merged_dir, 
                           outdir=demultiplexed_dir, 
                           check_reverse=TRUE, 
                           cutadapt_path=cutadapt_path, # can be omitted if CUTADAPT is in the PATH
                           vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
                           compress=FALSE)
```

**Dereplicate**

The fasta files produced by `SortReads` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?Dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r depelicate3}
outfile <- file.path(outdir, "1_before_filter.csv")
read_count_df <- Dereplicate(sampleinfo_df, 
                             dir=demultiplexed_dir, 
                             outfile=outfile)
```
