---
title: "From fastq to data frame - v1.0.4"
auhor: Emese Meglécz
date:  "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{From fastq to data frame - v1.0.4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


According to your wet lab and sequencing protocol, each FASTQ file may contain one or more [sample-replicates](tutorial-vtamr-pipeline.html#glossary). In addition, sequences may include [tags](tutorial-vtamr-pipeline.html#glossary) 
(used for [demultiplexing](tutorial-vtamr-pipeline.html#glossary)) 
as well as primer sequences.

In the following sections, we present **three common scenarios** for generating the [read_count_df](tutorial-vtamr-pipeline.html#read_count_df) data frame, 
which serves as the input for the filtering steps.

# Set up

*Load library*
```{r setup}
library(vtamR)
```

*Set path to third party programs*
```{r set_path_win}
# Example for Windows
cutadapt_path <- "C:/Users/Public/cutadapt"
vsearch_path <- "C:/Users/Public/vsearch-2.23.0-win-x86_64/bin/vsearch"
pigz_path <- "C:/Users/Public/pigz-win32/pigz" # optional to speed file compression
```

```{r set_path, echo=FALSE}
# Example for Windows
cutadapt_path <- "cutadapt"
vsearch_path <- "vsearch"
pigz_path <- "pigz" # optional to speed file compression
```

```{r set_path_linux}
#  Example for Linux
cutadapt_path <- "~/miniconda3/envs/vtam/bin/cutadapt" # v3.4
vsearch_path <- "~/miniconda3/envs/vtam/bin/vsearch" # v2.15.1
pigz_path <- "~/miniconda3/envs/vtam/bin/pigz" # optional to speed file compression
```

**Adapt the path** to third party programs according to your installation
(See [Installation](installation.html)).

If third party programs are in your `PATH` 
(See [Installation](installation.html#install-third-party-programs)), simply omit
the `cutadapt_path`, `pigz_path`, `vsearch_path` argument when calling the `vtamR` functions.

`compress = FALSE` In the tutorial we will use uncompressed files, but see
[To compress or not to compress](tutorial-vtamr-pipeline.html#to-compress-or-not-to-compress)
for the best strategy for you. 
See also [Files to keep](tutorial-vtamr-pipeline.html#files-to-keep) on how to handle the 
large intermediate files that should be compressed or deleted once the analyses are complete.

# Case 1 - One sample per fastq - no tag - no primer

In this scenario, each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads have been trimmed from all artificial add-ons, such as
adapters, tags, indices and also from primers.

Read pairs should be **quality filtered**, **merged** and written to 
**fasta format**.
This can be done by the `merge_fastq_pairs` function. 

See the help (`?merge_fastq_pairs`) for setting the 
correct parameters for quality filtering.

## Set Input

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): is either a csv file, or a data frame. The key 
information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged. 
The `tag_fw`, `primer_fw`, `tag_rv`, `primer_rv` are irrelevant in this case, 
just fill them with `NA`.
* `fastq_dir`: Directory containing the input fastq files.
* [sampleinfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `merge_fastq_pairs`. It is the updated 
version of fastqinfo, where fastq file names have been replaced by fasta file 
names and the read counts are included for each file. 
* `outdir`: Name of the output directory.


The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input1}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo1.csv", package = "vtamR")

outdir <- "vtamR_demo_case1"
merged_dir <- file.path(outdir, "merged")
```

## Merge Fastq File Pairs and Quality Filter Reads

```{r merge1}
sampleinfo_df <- merge_fastq_pairs(
  fastqinfo, 
  fastq_dir=fastq_dir, 
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  outdir=merged_dir,
  compress=FALSE
  )
```


## Dereplicate reads

The fasta files produced by `merge_fastq_pairs` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r derelicate1}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- dereplicate(
  sampleinfo_df, 
  dir=merged_dir, 
  outfile=outfile
  )
```


# Case 2 - One Sample per Fastq - Primer - No Tag

This is one of the most frequent case. 
Each pair of fasta files correspond to a sample (or
a replicate of a sample if you have replicates), 
so no demultiplexing is necessary.

The reads has been trimmed from all artificial add-ons, such as
adapters, tags, BUT they still have the primers.

Read pairs should be **quality filtered**, **merged** and written to **fasta format** by 
`merge_fastq_pairs` function as in the previous section.

Then the `trim_primer` function will **trim the primers** from the reads. 
See the help (`?trim_primer`) for setting the correct parameters for primer trimming.

## Set Input

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame.  
The key information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged. 
The `primer_fw`, `primer_rv` columns are irrelevant in this case, just fill them with `NA`.
* `fastq_dir`: Directory containing the input fastq files.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): is the output of `merge_fastq_pairs`. 
It is the updated version of `fastqinfo`, where fastq file names have been 
replaced by fasta file names containing the merged sequences.
* `fasta_dir`: Directory containing the input fasta files for `trim_primer`. 
This directory is created by `merge_fastq_pairs`.
* If `check_reverse` is TRUE, `trim_primer` checks the reverse complementary strand as well.
* `sampleinfo_df`: is  updated version of fastainfo. 
This data frame and the files listed in it are the input for `dereplicate`.
* `outdir`: Name of the output directory. 

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input2}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo2.csv", package = "vtamR")

outdir <- "vtamR_demo_case2"
merged_dir <- file.path(outdir, "merged")
```

## Merge Fastq File Pairs and Quality Filter Reads

```{r merge2}
fastainfo_df <- merge_fastq_pairs(
  fastqinfo, 
  fastq_dir=fastq_dir, 
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  outdir=merged_dir,
  compress=FALSE
  )
```

## Trim Primers

```{r primer-trim}

demultiplexed_dir <- file.path(outdir, "demultiplexed")
sampleinfo_df <- trim_primers(
  fastainfo_df, 
  fasta_dir=merged_dir, 
  outdir=demultiplexed_dir, 
  cutadapt_path=cutadapt_path, # can be omitted if CUTADAPT is in the PATH
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  check_reverse=T,
  primer_to_end=F,
  compress=FALSE
  )
```

## Dereplicate

The fasta files produced by `TrimPrimer` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?Dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r derelicate2}
outfile <- file.path(outdir, "1_before_filter.csv")

read_count_df <- dereplicate(
  sampleinfo_df, 
  dir=demultiplexed_dir, 
  outfile=outfile
  )
```

# Case 3 - Several Samples per Fastq - Tags - Primers

In this case, one pair of fastq files contains reads from multiples samples or
sample-replicates, so it is necessary to demultiplex them, and trim from 
tags and primers.

## Case 3a Merge - demultiplex

Read pairs should be **quality filtered**, **merged** and written to **fasta format** as 
in the previous sections.

Then the `demultiplex_and_trim` function will 
**[demultiplex](tutorial-vtamr-pipeline.html#glossary)** the fasta 
files according to the **[tag](tutorial-vtamr-pipeline.html#glossary)** combinations and **[trim](tutorial-vtamr-pipeline.html#glossary)** 
the primers from the reads. 

See the help (`?demultiplex_and_trim`) for setting the 
correct parameters for demultiplexing and primer trimming:

### Set Input

* [fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo): Either a csv file, or a data frame. 
The key information for `merge_fastq_pairs` is the list of the fastq file pairs that should be merged.
* `fastq_dir`: Directory containing the input fastq files.
* [fastainfo_df](tutorial-vtamr-pipeline.html#fastainfo): Output of `merge_fastq_pairs`. 
It is the updated version of `fastqinfo`, where fastq file names have been 
replaced by fasta file names.
* `fasta_dir`: Directory containing the input fasta files for `demultiplex_and_trim`. 
This directory is created by `merge_fastq_pairs`.
* If `check_reverse` is TRUE, `demultiplex_and_trim` checks the reverse complementary stand as well.
* `sampleinfo_df`: Updated version of `fastainfo`. 
This data frame and the files listed in it are the input of the `dereplicate`.
* `outdir`: Name of the output directory. 

The demo files below are included with the `vtamR` package, which is why we
use `system.file()` to access them in this tutorial. 
When using your own data, simply provide the file and directory names (e.g. `~/vtamR/fastq`).
Make sure there is **no space in the path and file names**.

```{r set-input3}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package = "vtamR")

outdir <- "vtamR_demo_case3a"
merged_dir <- file.path(outdir, "merged")
demultiplexed_dir <- file.path(outdir, "demultiplexed")
```

### Merge Fastq File Pairs and Quality Filter Reads

```{r merge3}
fastainfo_df <- merge_fastq_pairs(
  fastqinfo, 
  fastq_dir=fastq_dir, 
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  outdir=merged_dir,
  compress=FALSE
  )
```

### Demultiplex, Trim off Tags and Primers**
```{r demultiplex3}
sampleinfo_df <- demultiplex_and_trim(
  fastainfo_df, 
  fasta_dir=merged_dir, 
  outdir=demultiplexed_dir, 
  check_reverse=TRUE, 
  cutadapt_path=cutadapt_path, # can be omitted if CUTADAPT is in the PATH
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  compress=FALSE
  )
```

### Dereplicate

The fasta files produced by `demultiplex_and_trim` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r depelicate3}
outfile <- file.path(outdir, "1_before_filter.csv")
read_count_df <- dereplicate(
  sampleinfo_df, 
  dir=demultiplexed_dir, 
  outfile=outfile
  )
```


Here's a polished version with corrected grammar, clearer wording, and improved consistency:

## Case 3b: Demultiplex First, Then Merge

When submitting FASTQ files to the NCBI Sequence Read Archive (SRA), each file 
must contain reads from a single sample-replicate combination and must be trimmed 
of tags/adapters and primers. Therefore, instead of merging reads first and then 
demultiplexing them, as described in the previous section, the order of these 
steps can be reversed.

FASTQ file pairs can be demultiplexed and trimmed of tags and primers using 
the `demultiplex_fastq_pairs` function. This procedure produces FASTQ file pairs 
that can be submitted directly to the SRA.

The demultiplexed FASTQ files should then be merged and converted to FASTA format 
before proceeding with the filtering steps.

See the help page (`?demultiplex_fastq_pairs`) for details on setting the 
appropriate parameters for demultiplexing and primer trimming.

### Input Parameters

* **`fastqinfo`** ([fastqinfo](tutorial-vtamr-pipeline.html#fastqinfo)): 
Either a CSV file or a data frame. For `demultiplex_fastq_pairs`, the key 
information includes the list of FASTQ file pairs to be demultiplexed,
as well as the primer and tag sequences associated with each sample-replicate combination.
* **`fastq_dir`**: Directory containing the input FASTQ files.
* If **`check_reverse = TRUE`**, `demultiplex_fastq_pairs` can check the reverse-complement strand.
* **`sampleinfo_df`**: An updated version of `fastqinfo`. This data frame, 
together with the files listed in it, serves as input for the `dereplicate` step.
* **`outdir`**: Name of the output directory.

The example files used below are included with the `vtamR` package, 
which is why we use `system.file()` to access them in this tutorial.
When using your own data, simply provide the appropriate file and directory 
paths (e.g., `~/vtamR/fastq`). 
Make sure that there are **no spaces in file or directory names**.

```{r set-input4}
fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <-  system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package = "vtamR")

outdir <- "vtamR_demo_case3b"
demultiplexed_dir <- file.path(outdir, "demultiplexed")
merged_dir <- file.path(outdir, "merged")
```


### Demultiplex, Trim off Tags and Primers and keep fastq format
```{r demultiplex4}
fastqinfo_delultiplexed <- demultiplex_fastq_pairs(
  fastqinfo = fastqinfo, 
  fastq_dir = fastq_dir, 
  outdir = demultiplexed_dir, 
  cutadapt_path=cutadapt_path,
  primer_to_end = FALSE,
  check_reverse=TRUE,
  compress=FALSE,
  quiet=TRUE
  )
```

### Merge Fastq File Pairs and Quality Filter Reads

```{r merge4}
sampleinfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo_delultiplexed, 
  fastq_dir= demultiplexed_dir, 
  vsearch_path=vsearch_path, # can be omitted if VSEARCH is in the PATH
  outdir=merged_dir,
  compress=FALSE
  )
```


### Dereplicate

The fasta files produced by `merge_fastq_pairs` can be read to a data frame and be 
[dereplicated](tutorial-vtamr-pipeline.html#glossary) by the `Dereplicate` function.
See the help (`?dereplicate`) and [tutorial](tutorial-vtamr-pipeline.html#dereplicate) more
more information.

```{r depelicate4}
outfile <- file.path(outdir, "1_before_filter.csv")
read_count_df <- dereplicate(
  sampleinfo_df, 
  dir=merged_dir, 
  outfile=outfile
  )
```
