---
title: "Install vtamR and Third Party Programs - v0.3.0"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Install vtamR and Third Party Programs - v0.3.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

## Implementation

The heavy lifting of sequence analyses are done by the following third party 
programs. They should be installed on your computer:

* BLAST([Altschul et al., 1990](https://pubmed.ncbi.nlm.nih.gov/2231712/)) (v2.11.0+)
* vsearch ([Rognes et al., 2016](https://peerj.com/articles/2584/)) (v2.7.0)
* cutadapt([Martin, 2011](https://journal.embnet.org/index.php/embnetjournal/article/view/200/479)) (v4.0)
* swarm ([Mahé et al., 2015](https://peerj.com/articles/1420/)) (v2.1.12)

`vtamR` has been tested using the above mentioned versions, 
but it should work with later versions.
`vtamR` was tested on **Windows** and **Linux**, and should work in all 
operating systems.

 - Help is available for all `vtamR` functions via `?function_name`.
 - Most functions accept either a **data frame** or a **CSV file** as input, 
 and return a data frame (optionally saved to a CSV file if a filename is provided).


## Install vtamR

`pkg_install` will handle conveniently the R dependencies of `vtamR`. 
```{r install_github_pak, eval=FALSE}
if(!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
pak::pkg_install("meglecz/vtamR@develop", dependencies=TRUE)
```

Alternatively, you can use devtools, but you need to install Bioconductor packages manually
if you intend to use `TaxAssignRDP` function.

```{r install_github_devtools, eval=FALSE}
# Install vtamR and all CRAN dependencies
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("meglecz/vtamR@develop", dependencies = TRUE)

# Install Bioconductor suggested packages manually if needed (only for TaxAssignRDP)
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("rRDP", quietly = TRUE)) BiocManager::install("rRDP")
if (!requireNamespace("rRDPData", quietly = TRUE)) BiocManager::install("rRDPData")
if (!requireNamespace("Biostrings", quietly = TRUE)) BiocManager::install("Biostrings")
```

Read the **[Manual](https://people.imbe.fr/~emeglecz/vtamR)** 
on how to construct a full metabarcoding pipeline.

## Install third party programs

### Option 1: Install third party programs using conda environment
**NOTE:** Should be updated to include pigz and tested

**In R**, find the full path to the YAML environment file:

```{R environment_vtamR, eval=FALSE}
environment_vtamR_yml_path <- system.file("environment_vtamR.yml", package = "vtamR")
cat(environment_vtamR_yml_path)
```

**In a terminal**, replace PATH_TO_YML with the path you just printed:

```{bash instal-conda, eval=FALSE}
conda env create -f PATH_TO_YML
conda activate vtamRenv
```

This creates a Conda environment `vtamRenv` with all required tools.

If using **RStudio**, you must manually set the paths to these tools in 
your R session or scripts.
See the [Set up section of the tutorial](tutorial-vtamr-pipeline.html#setup) 
for details.

Get paths to third-party programs:
```{bash access-path, eval=FALSE}
conda activate vtamRenv

which vsearch
which blastn
which cutadapt
which swarm
```

Check the version of third-party programs:
```{bash check-version, eval=FALSE}
vsearch --version # >= v2.7.0
blastn -v # >= v2.11.0+
cutadapt --version  # >= v4.0
swarm --version # >= v2.1.12
```

### Option 2: Install third party programs manually

#### Linux

* `vsearch` [https://github.com/torognes/vsearch](https://github.com/torognes/vsearch)
* `BLAST` [https://www.ncbi.nlm.nih.gov/books/NBK52640/](https://www.ncbi.nlm.nih.gov/books/NBK52640/)
* `cutadapt` [https://cutadapt.readthedocs.io/en/stable/installation.html](https://cutadapt.readthedocs.io/en/stable/installation.html)
* `swarm` [https://github.com/torognes/swarm](https://github.com/torognes/swarm)
* `pigz` (optional) [https://github.com/madler/pigz](https://github.com/madler/pigz)

**Add the executables to your PATH** to avoid specifying their full path each 
time you run a function that uses them.

#### Windows

Download binaries and save it a convenient place on your computer 
(path without space, e.g. `C:/Users/Public/`)

**vsearch**

* Download binaries from https://github.com/torognes/vsearch/releases/tag/v2.23.0
* Decompress the zip file
* The executable are found in `vsearch-x.xx.x-win-x86_64/bin/`

**cutadapt**

* Download `cutadapt.exe` from 
[https://github.com/marcelm/cutadapt/releases](https://github.com/marcelm/cutadapt/releases)

**swarm**

* Download binaries from 
[https://github.com/torognes/swarm/releases](https://github.com/torognes/swarm/releases)
* Decompress the zip file
* The executable is in the `swarm-x.x.x-win-x86_64/bin` directory

**BLAST**

Detailed instructions:
https://www.ncbi.nlm.nih.gov/books/NBK52637/

* Download the installer (ncbi-blast-x.xx.x+-win64.exe) from:
https://ftp.ncbi.nlm.nih.gov/blast/executables/LATEST/
* Double-click the .exe file, accept the license agreement, 
and keep the suggested installation path.
This will automatically add BLAST to your system PATH, 
so you won’t need to specify its location when using it.
* The executables will be installed in:
`C:\Program Files\NCBI\blast-x.xx.x+\bin`

**pigz** (Optional)

* Download [pigz-win32.zip](https://sourceforge.net/projects/pigz-for-windows/files/)
* Decompress the zip file
* The executable (pigz.exe) is in the pigz-win32 directory


**Add executables to your PATH**

This step is optional. If third-party executables are included in your 
system PATH, you won’t need to manually specify their locations when running 
functions that use them.

If you choose **not** to add them to your PATH, remember to specify the full 
path to each program when calling those functions.


## TaxAssign reference data base

A ready-to-use COI database is available from OSF [OSF](https://osf.io/vrfwz/), 
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).

Download and extract using:
```{r, dowload-coinr, eval=FALSE}
library(vtamR)

download_osf(filename = "COInr_for_vtam_2025_05_23_dbV5.tar.gz",
    url = "https://osf.io/download/jyhz6/",
    dest_dir = "COInr_db",
    untar = TRUE,
    quiet = FALSE
    )
```

To check for newer versions:

- Go to: [https://osf.io/vrfwz/files/osfstorage](https://osf.io/vrfwz/files/osfstorage)
- In **Files** open the the **dbV5** folder.
- Choose your version → Click on the tree-dot menu → Right-click Download → Copy Link Address.


Alternatively, you can also create a custom version from the **TSV format** database available at
[Zenodo](https://zenodo.org/records/15515860) and make a custom version using 
[mkCOInr](https://github.com/meglecz/mkCOInr)
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).

**For other markers** you will need a [database formatted to BLAST](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments), 
containing taxIDs and a 
[taxonomy file](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments).
If all sequences are extracted from NCBI-nt. The taxonomy file provided in 
[OSF](https://osf.io/vrfwz/) can be used without modification.

