---
title: "Install vtamR and Third Party Programs - v1.0.0"
auhor: Emese Meglécz
date:  "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Install vtamR and Third Party Programs - v1.0.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

## Implementation

The heavy lifting of sequence analyses are done by the following third party 
programs. They should be installed on your computer:

* BLAST([Altschul et al., 1990](https://pubmed.ncbi.nlm.nih.gov/2231712/)) (v2.11.0+)
* VSEARCH ([Rognes et al., 2016](https://peerj.com/articles/2584/)) (v2.7.0)
* CUTADAPT([Martin, 2011](https://journal.embnet.org/index.php/embnetjournal/article/view/200/479)) (v4.0)
* SWARM ([Mahé et al., 2015](https://peerj.com/articles/1420/)) (v2.1.12)
* PIGZ (optional)

`vtamR` has been tested using the above mentioned versions, 
but it should work with later versions.

`vtamR` was tested on **Windows** and **Linux**, and should work in all 
operating systems.


## Install vtamR

Install `vtamR` and its CRAN dependencies by `pak`.
```{r install_github_pak, eval=FALSE}
if(!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
pak::pkg_install("meglecz/vtamR", dependencies=TRUE)
```

Alternatively, you can use `devtools`.
```{r install_github_devtools, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("meglecz/vtamR", dependencies = TRUE)
```

Install Bioconductor packages manually
if you intend to use `assign_taxonomy_rdp` function.
```{r install_bioc, eval=FALSE}
# Install Bioconductor suggested packages manually if needed (only for assign_taxonomy_ltg)
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("rRDP", quietly = TRUE)) BiocManager::install("rRDP")
if (!requireNamespace("rRDPData", quietly = TRUE)) BiocManager::install("rRDPData")
if (!requireNamespace("Biostrings", quietly = TRUE)) BiocManager::install("Biostrings")
```

Read the **[Manual](https://people.imbe.fr/~emeglecz/vtamR)** 
on how to construct a full metabarcoding pipeline.

## Install third party programs

### Linux

* `VSEARCH` [https://github.com/torognes/vsearch](https://github.com/torognes/vsearch)
* `BLAST` [https://www.ncbi.nlm.nih.gov/books/NBK52640/](https://www.ncbi.nlm.nih.gov/books/NBK52640/)
* `CUTADAPT` [https://cutadapt.readthedocs.io/en/stable/installation.html](https://cutadapt.readthedocs.io/en/stable/installation.html)
* `SWARM` [https://github.com/torognes/swarm](https://github.com/torognes/swarm)
* `PIGZ` (optional) [https://github.com/madler/pigz](https://github.com/madler/pigz)


**Add the executables to your PATH** 

This step is optional. If third-party executables are included in your `PATH`, 
you won’t need to manually specify their locations when running functions that use them.

If you prefer **not** to modify your `PATH`, be sure to provide the 
full path to each program using the corresponding `vtamR` function argument 
(e.g., `vsearch_path`).


1. Edit (or create) your `~/.Renviron` file

```bash
nano ~/.Renviron
```

2. Add the directory to the PATH

```bash
PATH="/home/user/tools/myprogram/bin:${PATH}"
```
**Note:** The PATH must contain a **directory**, not an executable file.
For example, use: `/home/username/miniconda3/envs/vtamr/bin`
and **not**: `/home/username/miniconda3/envs/vtamr/bin/cutadapt`

3. Restart R or RStudio

The new PATH is loaded at startup.

4. Verify that R sees the correct executables

```r
system("cutadapt --version")
system("swarm --version")
system("vsearch --version")
system("blastn -version")
system("pigz --version")
```

### Windows

Download binaries and save them to a convenient place on your computer 
(path without space, e.g. `C:/Users/Public/`)

**VSEARCH**

* Download binaries from https://github.com/torognes/vsearch/releases/tag/v2.23.0
* Decompress the zip file
* The executable are found in `vsearch-x.xx.x-win-x86_64/bin/`

**CUTADAPT**

* Download `cutadapt.exe` from 
[https://github.com/marcelm/cutadapt/releases](https://github.com/marcelm/cutadapt/releases)

**SWARM**

* Download binaries from 
[https://github.com/torognes/swarm/releases](https://github.com/torognes/swarm/releases)
* Decompress the zip file
* The executable is in the `swarm-x.x.x-win-x86_64/bin` directory

**BLAST**

* Download the installer (ncbi-blast-x.xx.x+-win64.exe) from:
https://ftp.ncbi.nlm.nih.gov/blast/executables/LATEST/
* Double-click the .exe file, accept the license agreement, 
and keep the suggested installation path.
This will automatically add BLAST to your system PATH, 
so you won’t need to specify its location when using it.
* The executables will be installed in:
`C:\Program Files\NCBI\blast-x.xx.x+\bin`

**VSEARCH** (Optional)

* Download [pigz-win32.zip](https://sourceforge.net/projects/pigz-for-windows/files/)
* Decompress the zip file
* The executable (pigz.exe) is in the pigz-win32 directory

**Add executables to your PATH**

This step is optional. If third-party executables are included in your 
system PATH, you won’t need to manually specify their locations when running 
functions that use them.

If you prefer **not** to add them, make sure to provide the full path to each program through the appropriate `vtamR` function argument (e.g., `vsearch_path`).

You can use either **GUI** or **PowerShell** to add executables to your PATH.

1. Using the Windows GUI (System Settings)

* Press **Win + R**, type `sysdm.cpl`, and press **Enter**.
   *(Opens System Properties.)*
* Go to the **Advanced** tab.
* Click **Environment Variables…**
* In the **System variables** section, find and select **Path**.
* Click **Edit…**
* Click **New**, then paste the directory you want to add (for example:
   `C:\Users\YourName\Tools\bin`)
* Click **OK** on all windows to save.

2. Using PowerShell

* Press **Win + R**, type `powershell`, and press **Enter**.
   *(Opens Powershell)*

```{}
setx PATH "$($env:Path);C:\My\New\Folder\vsearch:C:\My\New\Folder\cutadapt;C:\My\New\Folder\pigz;C:\My\New\Folder\swarm"
```

* Close de Powershell, since the change applies only to **new** terminals, not the current one.


**Verify Program Versions**

Open a Command Prompt on Windows (**Win + R**, type `cmd`)

```bash
cutadapt --version
swarm --version
vsearch --version
pigz --version
blastn -version
```

**Note**
*These commands will display the version of each program **if the executables are in your PATH**.
Otherwise, don’t forget to provide the full path to the executables.*

Example:

```{}
C:/Users/Public/vsearch-x.xx.x-win-x86_64/bin/vsearch --version
```

## TaxAssign reference data base

### assign_taxonomy_ltg

For `assign_taxonomy_ltg` (BLAST based Lowest Common Ancestor method),
a ready-to-use COI database is available from OSF [OSF](https://osf.io/vrfwz/), 
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).

Download and extract using:
```{r, dowload-coinr, eval=FALSE}
library(vtamR)

download_osf(filename = "COInr_for_vtam_2025_05_23_dbV5.tar.gz",
    url = "https://osf.io/download/jyhz6/",
    dest_dir = "COInr_db",
    untar = TRUE,
    quiet = FALSE
    )
```

To check for newer versions:

- Go to: [https://osf.io/vrfwz/files/osfstorage](https://osf.io/vrfwz/files/osfstorage)
- In **Files** open the the **dbV5** folder.
- Choose your version → Click on the tree-dot menu → Right-click Download → Copy Link Address.


Alternatively, you can also create a custom version from the **TSV format** database available at
[Zenodo](https://zenodo.org/records/15515860) and make a custom version using 
[mkCOInr](https://github.com/meglecz/mkCOInr)
([Meglécz, 2023](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13756)).


**For other markers** you will need a [database formatted to BLAST](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments-with-assign_taxonomy_ltg), 
containing taxIDs and a 
[taxonomy file](tutorial-vtamr-pipeline.html#reference-database-for-taxonomic-assignments-with-assign_taxonomy_ltg).
If all sequences are extracted from NCBI-nt, the taxonomy file provided in 
[OSF](https://osf.io/vrfwz/) can be used without modification.


### assign_taxonomy_rdp

`assign_taxonomy_rdp` function uses de RDP algorithm to assign bacterial 16S sequences.
The database is provided in the `rRDPData` R data package.

It is also possible to provide a trained classifier object created 
with `trainRDP()`.




