---
title: "How to make a mock_composition file - v1.0.3"
auhor: Emese Meglécz
date:  "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to make a mock_composition file - v1.0.3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=TRUE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


# What it is and when to use it

The [mock_composition](tutorial-vtamr-pipeline.html#mock_composition) file
defines which ASVs are expected in each mock sample and is a key input for the analysis.

It is provided as a CSV file with the following columns:

* sample: name of the [mock](tutorial-vtamr-pipeline.html#glossary) sample
* action:

  * `keep`: ASV expected in the mock that should be retained in the dataset
  * `tolerate`: ASV that may be present in the mock but is not critical to retain (e.g. poorly amplified organism)
* [asv](tutorial-vtamr-pipeline.html#glossary): ASV sequence
* taxon: optional; organism name
* [asv_id](tutorial-vtamr-pipeline.html#glossary): optional; if both `asv` and `asv_id` are provided and conflict, `asv_id` is ignored

The [known_occurrences](tutorial-vtamr-pipeline.html#known_occurrences) file
is required to run the following functions:

* `suggest_pcr_error_cutoff`
* `suggest_sample_cutoff`
* `suggest_variant_readcount_cutoffs`
* `compute_asv_specific_cutoff`
* `classify_control_occurrences`

These functions are used to:

* assess filtering performance in terms of precision (TP / (TP + FP)) 
and sensitivity (TP / (TP + FN)) — via `classify_control_occurrences`
* identify optimal parameter values for several filters (`suggest_pcr_error_cutoff`, 
`suggest_sample_cutoff`, `suggest_variant_readcount_cutoffs`, `compute_asv_specific_cutoff`)

The [mock_composition](tutorial-vtamr-pipeline.html#mock_composition) file is 
also useful—though not required—for the `write_asv_table` function, 
if you want to include an output column that highlights expected occurrences in 
each mock sample.


# Constructing the `mock_composition` File

I suggest two different methods to construct the `mock_composition` file:

1. Using the `match_variants_to_mock_species` function
2. Prefiltering the dataset and manually selecting the expected ASVs

## 1. Using `match_variants_to_mock_species`

This function offers a quick and convenient way to generate a 
`mock_composition` file.
However, it may occasionally miss some expected occurrences. 
In such cases, you can manually retrieve the expected ASVs from a 
prefiltered dataset (see below).

### Prerequisites

* **Reference sequences**

  Prepare one reference sequence for each species expected in the mock samples.
  Each sequence should span at least 70% of the region targeted by the primers.
  It can be longer or slightly shorter than the ASV and may include minor 
  mismatches relative to the true sequence.

  Reference sequences must be provided in FASTA format, with headers including a valid NCBI taxonomic identifier:

  ```
  >SequenceName taxID=12345
  ```

  The `taxID` must correspond to an entry in the
  [NCBI Taxonomy database](https://www.ncbi.nlm.nih.gov/taxonomy).

* **Taxonomy file**

  A taxonomy file is required.
  The file distributed with the **COInr** database is appropriate, even for non-COI markers
  (see [TaxAssign reference data base](installation.html#taxassign-reference-data-base)).

* **Read count data**

  Provide a `read_count_df` data frame containing read counts for each mock sample.
  For better performance, it is recommended to use data that has already undergone 
  initial filtering to remove artefactual ASVs and substantially reduce dataset size
  (e.g. after Denoising with SWARM, `LFNglobalReadCount`, `FilterIndel`,
  `FilterCodonStop`, `FilterExternalContaminant`, and `FilterChimera`).

  Although this improves speed, the function can also be run on unfiltered data.

### Function Overview

The `match_variants_to_mock_species` function performs the following steps:

1. **Builds a small BLAST database** from reference sequences representing 
expected species or closely related taxa.
2. **Assigns taxonomy** to all ASVs detected in mock samples using this 
custom database. Because the database is small, this step is fast.
3. **Selects the most abundant ASV** for each taxon.
4. **Generates a `mock_composition` template file**, which should be reviewed 
and edited if necessary.

### Run `match_variants_to_mock_species`

We will use files generated in the first part of the
[Tutorial](tutorial-vtamr-pipeline.html) (up to
[filter_replicate](tutorial-vtamr-pipeline.html#filter-replicate)).

* Demo files are included in the `vtamR` package (hence the use of `system.file()`).
  When using your own data, simply replace these with your file paths.
* `read_count_file` corresponds to the output of `filter_replicate` from the
  [Tutorial](tutorial-vtamr-pipeline.html).
* `blast_db` and `taxonomy` are configured as described in the
  [Tutorial](tutorial-vtamr-pipeline.html#database-for-taxonomic-assignment).

```{r set_db, eval=TRUE}
library(vtamR)

read_count_file <- system.file("extdata/demo/8_filter_replicate.csv", package = "vtamR")
reference_mock_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package = "vtamR")
sampleinfo <- system.file("extdata/demo/sampleinfo_mfzr_plate1.csv", package = "vtamR")
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_path <- "blastn" # Adapt this if BLAST is not in your PATH

outdir_mock <- "mock_composition"

mock_template <- match_variants_to_mock_species(
  read_count=read_count_file,
  fas=reference_mock_fasta,
  taxonomy=taxonomy,
  sampleinfo = sampleinfo,
  outdir= outdir_mock,
  blast_path=blast_path
  )
```

**Note**: If BLAST is available in your PATH (see [Installation](installation.html)),
you can omit the `blast_path` argument.

### Output

The main output file is **`mock_composition_template_to_check.csv`**,
which serves as a template for the final `mock_composition` file.

```{r show_results, eval=TRUE}
mock_composition <- file.path(outdir_mock, "mock_composition_template_to_check.csv")
mock_composition_df <- read.csv(mock_composition)

knitr::kable(mock_composition_df, format = "markdown")
```

It contains the most abundant sequence for each `ltg_name` identified 
during taxonomic assignment,
replicated across all mock samples.
If mock samples differ in composition, you should remove entries 
corresponding to taxa not expected in a given sample.

**Note:**
The file does **not** include species that did not show significant similarity to any ASV.
This can happen if:

* the species was not amplified in the mock samples, or
* the reference sequence in the custom database is incorrect.

Here is your text with **correct R Markdown internal link syntax preserved and consistent** (anchors cleaned + safer IDs where needed).


## 2. Prefiltering the dataset and manually picking the expected ASVs

The idea is to 

- Prefilter your data set
- Assign ASV to taxa
- Examine the ASV in the mock samples and their read counts and pick the correct sequences.

I suggest that you start by filtering/denoising your data set by using at least some of
the following functions as in the [Tutorial](tutorial-vtamr-pipeline.html). 
This will eliminate most of the erroneous ASV, so
it will be easier to identify the expected ASV from your mock samples. 

- [Denoising with SWARM](tutorial-vtamr-pipeline.html#denoising-with-swarm)
- [filter_asv_global](tutorial-vtamr-pipeline.html#filter_asv_global)
- [filter_indel](tutorial-vtamr-pipeline.html#filter_indel)
- [filter_stop_codon](tutorial-vtamr-pipeline.html#filter_stop_codon)
- [filter_chimera](tutorial-vtamr-pipeline.html#filter_chimera)

### Set parameters and access the demo files

We will use some of the files created by the first part of the 
[Tutorial](tutorial-vtamr-pipeline.html) (till the 
[filter_replicate](tutorial-vtamr-pipeline.html#filter-replicate))

- The demo files are included in the `vtamR` package, hence the use of 
`system.file()`. When using your own data just enter your file names.
- `read_count_file` is the output of `filter_replicate` of the
[Tutorial](tutorial-vtamr-pipeline.html#filter-replicate).
- The `blast_db` and `taxonomy` are set up as in the 
[Tutorial](tutorial-vtamr-pipeline.html#database-for-taxonomic-assignment)


```{r set_db2}
library(vtamR)
library(dplyr)

read_count_file <- system.file("extdata/demo/8_filter_replicate.csv", package = "vtamR")
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
sampleinfo <- system.file("extdata/demo/sampleinfo_mfzr_plate1.csv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")
blast_path <- "blastn" # Adapt this if BLAST is not in your PATH
```

Let's limit the analyses to the mock samples

```{r}
read_count_df <- read.csv(read_count_file)
sampleinfo_df <- read.csv(sampleinfo)

# select mock samples in sampleinfo
mock_samples <- sampleinfo_df %>%
  filter(sample_type == "mock")

# select mock samples from read_count_df
read_count_mock <- read_count_df %>%
  filter(sample %in% mock_samples$sample)
```

### Assign taxa to ASVs {#taxassign}

`assign_taxonomy_ltg` will assign all ASV in the input csv file or data frame (`read_count_file`).

See more details of taxonomic assignment [here](tutorial-vtamr-pipeline.html#taxonomic-assignment).

**Note**: If BLAST is in your PATH, you can omit the `blast_path` argument.

```{r}
asv_tax <- assign_taxonomy_ltg(
  asv = read_count_mock, 
  taxonomy = taxonomy, 
  blast_db = blast_db,
  quiet = TRUE,
  blast_path = blast_path
)
```

### Make an ASV table with taxonomic assignments {#print-output}

Make a data frame with ASVs and read counts in the
[wide format](tutorial-vtamr-pipeline.html#glossary)
and add their taxonomic assignment.

This format is easier to read for humans than `read_count_df`.

See details of `write_asv_table` [here](tutorial-vtamr-pipeline.html#print-output).

```{r}
asv_table_mock <- write_asv_table(
  read_count_mock, 
  sampleinfo = sampleinfo, 
  asv_tax = asv_tax,
  pool_replicates = TRUE
)
```

Sort the output by the taxon name and then by decreasing read count.

```{r}
asv_table_mock <- asv_table_mock %>%
  arrange(ltg_name, desc(tpos1))
```

Let's see the ASV present in `tpos1`.

```{r}
knitr::kable(asv_table_mock, format = "markdown")
```

In this mock sample, there should be the following 6 species:

* *Caenis pusilla*
* *Rheocricotopus*
* *Phoxinus phoxinus*
* *Hydropsyche pellucidula*
* *Synorthocladius semivirens*
* *Baetis rhodani*

We can see that in spite of all the filtering we have done so far, there are still
a lot of unexpected occurrences in this sample.

Most of them have low read counts and could be filtered out by
[Low Frequency Noise Filters](tutorial-vtamr-pipeline.html#glossary) such as 
`filter_occurrence_read_count`, `filter_occurrence_sample`, `filter_occurrence_variant`.


### Select the expected ASV and make mock_composition {#mock-composition}

You can now pick the correct sequences of the expected ASVs in each mock and make the
[mock_composition](tutorial-vtamr-pipeline.html#mock_composition) file.


