---
title: "How to make a mock_composition file - v0.3.1"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to make a mock_composition file - v0.3.1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo=TRUE,
  eval=TRUE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


## What is it and when to use it

The [mock_composition](tutorial-vtamr-pipeline.html#mock_composition) file
is essential to describe which ASVs are expected in each mock sample.

It is a CSV file with the following columns:

 * sample: Name of the [mock](tutorial-vtamr-pipeline.html#glossary) sample
 * action: 
      * `keep`: Expected ASV in the mock, that should be kept in the data set
      * `tolerate`: ASV that can be present in a mock, but it is not essential to keep it in the data set (e.g. badly amplified organism)
 * [asv](tutorial-vtamr-pipeline.html#glossary): sequence of the ASV
 * taxon: Optional; Name of the organism
 * [asv_id](tutorial-vtamr-pipeline.html#glossary): Optional; If there is a conflict between asv and asv_id, the asv_id is ignored

The [known_occurrences](tutorial-vtamr-pipeline.html#known_occurrences) file is 
essential for running the following functions, either to determine the 
optimal parameter values for the LFN filters or to evaluate the filtering 
procedure in terms of precision (TP / (TP + FP)) and sensitivity (TP / (TP + FN)).

* `OptimizePCRerror`
* `OptimizeLFNsampleReplicate`
* `OptimizeLFNreadCountLFNvariant`
* `MakeKnownOccurrences`
* `ASVspecificCutoff`

The [mock_composition](tutorial-vtamr-pipeline.html#mock_composition) is also useful, 
although not essential for the `WriteASVtable` function if you wish to add a column in 
the output to easily find expected occurrences in each mock sample.


## Constructing the `mock_composition` File

I suggest two different methods to construct the `mock_composition` file:

1. Using the `MakeMockCompositionLTG` function
2. Prefiltering the dataset and manually selecting the expected ASVs


### 1. Using `MakeMockCompositionLTG`

This function provides a convenient and fast way to build the `mock_composition` file.
However, some expected occurrences may occasionally be missed.
If that happens, you can manually select the expected ASVs from a prefiltered dataset 
(as descirbed later).

#### Prerequisites

* **Reference sequences**

  Collect a reference sequence for each species expected in the mock samples.
  Each reference should cover at least 70% of the region amplified by the primers.
  It may be longer or slightly shorter than the ASV and may differ 
  slightly from the exact expected sequence.

  The reference sequences must be in FASTA format, and the FASTA 
  headers should include a valid NCBI taxonomic identifier in the following format:

  ```
  >SequenceName taxID=12345
  ```

  The `taxID` must correspond to a valid entry in the
  [NCBI Taxonomy database](https://www.ncbi.nlm.nih.gov/taxonomy).

* **Taxonomy file**

  The function requires a taxonomy file.
  The file distributed with the **COInr** database is suitable for Eukaryotes, even
  if it is not a COI marker.
  (see [TaxAssign reference data base](installation.html#taxassign-reference-data-base)).

* **Read count data**

  Provide a `read_count_df` data frame containing read counts for each mock sample.
  It is recommended to use a dataset that has already undergone initial 
  filtering steps to remove artefactual ASVs and reduce drastically the number of 
  ASV (e.g., after Denoising with SWARM, `LFNglobalReadCount`, `FilterIndel`, 
  `FilterCodonStop`, `FilterExternalContaminant`, and `FilterChimera`).

#### Function Overview

The `MakeMockCompositionLTG` function performs the following steps:

1. **Creates a small BLAST database** from the reference sequences corresponding 
to the expected species or closely related taxa.
2. **Assigns taxonomy** to all ASVs detected in the mock samples using this custom database.
   Since the database is small, the assignment is fast. 
3. **Selects the most abundant ASV** for each taxon.
4. **Generates a `mock_composition` template file**, which should be reviewed 
and, if necessary, edited by the user.

#### Output

The main output file is **`mock_composition_template_to_check.csv`**.
This file serves as a template for the final `mock_composition` file.

It contains the most abundant sequence for each `ltg_name` identified in 
the taxonomic assignment output, repeated for each mock sample.
If different mock samples have distinct compositions, you should 
remove the lines corresponding to taxa that are not expected in a particular sample.

**Note:**
The file does **not** include sequences for species in the custom database that 
did not show significant similarity to any ASV.
This may occur if:

* the species was not amplified in the mock samples, or
* the reference sequence in the custom database is incorrect.

We will use some of the files created by the first part of the 
[Tutorial](tutorial-vtamr-pipeline.html) (Till the 
[FilterRenkonen](tutorial-vtamr-pipeline.html#filterrenkonen))

- The demo files are included in the `vtamR` package, hence the use of 
`system.file()`. When using your own data just enter your file names.
- `read_count_file` is the output of `FilterRenkonen` of the
[Tutorial](tutorial-vtamr-pipeline.html).
- The `blast_db` and `taxonomy` are set up as in the 
[Tutorial](tutorial-vtamr-pipeline.html#set-up)

```{r set_db, eval=FALSE}
library(vtamR)

read_count_file <- system.file("extdata/demo/7_FilterChimera.csv", package = "vtamR")
reference_mock_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package = "vtamR")
sampleinfo <- system.file("extdata/demo/sampleinfo.csv", package = "vtamR")
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_path <- "blastn" # Adapt this if BLAST is not in your PATH

outdir_mock <- "mock_composition"

mock_template <- MakeMockCompositionLTG(read_count=read_count_file,
                                        fas=reference_mock_fasta,
                                        taxonomy=taxonomy,
                                        sampleinfo = sampleinfo,
                                        outdir= outdir_mock;
                                        blast_path=blast_path)
```

**Note**: If BLAST is in your PATH (see [Installation](installation.html)), 
you you can omit the `blast_path` argument.

### 2. Prefiltering the dataset and manually picking the expected ASVs

The idea is to 

- Prefilter your data set
- Assign ASV to taxa
- Examine the ASV in the mock samples and their read counts and pick the correct sequences.

I suggest that you start by filtering/denoising your data set by using at least some of
the following functions as in the [Tutorial](tutorial-vtamr-pipeline.html). 
This will eliminate most of the erroneous ASV, so
it will be easier to identify the expected ASV from your mock samples. 

- [Denoising by SWARM](tutorial-vtamr-pipeline.html#denoising-by-swarm)
- [LFNglobalReadCount](tutorial-vtamr-pipeline.html#lfnglobalreadcount)
- [FilterIndel](tutorial-vtamr-pipeline.html#filterindel)
- [FilterCodonStop](tutorial-vtamr-pipeline.html#filtercodonstop)
- [FilterChimera](tutorial-vtamr-pipeline.html#filterchimera)

### Set parametres and access the demo files

We will use some of the files created by the first part of the 
[Tutorial](tutorial-vtamr-pipeline.html) (Till the 
[FilterChimera](tutorial-vtamr-pipeline.html#filterchimera))

- The demo files are included in the `vtamR` package, hence the use of 
`system.file()`. When using your own data just enter your file names.
- `read_count_file` is the output of `FilterRenkonen` of the
[Tutorial](tutorial-vtamr-pipeline.html).
- The `blast_db` and `taxonomy` are set up as in the 
[Tutorial](tutorial-vtamr-pipeline.html#set-up)


```{r set_db2}
library(vtamR)
library(dplyr)

read_count_file <- system.file("extdata/demo/7_FilterChimera.csv", package = "vtamR")
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
sampleinfo <- system.file("extdata/demo/sampleinfo.csv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")
blast_path <- "blastn" # Adapt this if BLAST is not in your PATH
```

Let's limit the analyses to the mock samples

```{r}

read_count_df <- read.csv(read_count_file)
sampleinfo_df <- read.csv(sampleinfo)

# select mock samples in sampleinfo
mock_samples <- sampleinfo_df %>%
  filter(sample_type == "mock")
# select mock samples from read_count_df
read_count_mock <- read_count_df %>%
  filter(sample %in% mock_samples$sample)

```

### Assign taxa to ASVs

`TaxAssignLTG` will assign all ASV in the input csv file or data frame (`read_count_file`).

See more details of taxonomic assignment [here](tutorial-vtamr-pipeline.html#taxassign).

**Note**: If BLAST is in your PATH (see [Installation](installation.html)), 
you you can omit the `blast_path` argument.

```{r}
asv_tax <- TaxAssignLTG(asv=read_count_mock, 
                     taxonomy=taxonomy, 
                     blast_db=blast_db,
                     quiet=TRUE,
                     blast_path = blast_path 
                     )
```


### Make an ASV table with taxonomic assignments

Make a data frame with ASVs and read counts in the 
[wide format](tutorial-vtamr-pipeline.html#glossary) 
and add their taxonomic assignment. 
This format is easier to read for humans, than the `read_count_df`.

See details of `WriteASVtable` [here](tutorial-vtamr-pipeline.html#print-output).

```{r}

asv_table_mock <- WriteASVtable(read_count_mock, 
                               sampleinfo=sampleinfo, 
                               asv_tax=asv_tax,
                               pool_replicates=TRUE)
```

Sort the output by the taxon name and then by decreasing read count.
```{r}
asv_table_mock <- asv_table_mock %>%
  arrange(ltg_name, desc(tpos1))
```

Let's see the ASV present in `tpos1`.

```{r}
knitr::kable(asv_table_mock, format = "markdown")
```

In this mock sample, there should be the following 6 species:

- *Caenis pusilla*
- *Rheocricotopus*
- *Phoxinus phoxinus*
- *Hydropsyche pellucidula*
- *Synorthocladius semivirens*
- *Baetis rhodani*

We can see that in spite of all the filtering we have done so far, there are still
a lot of unexpected occurrences in this sample. 
Most of them have low read counts and could be filtered out by 
[Low Frequency Noise Filters](tutorial-vtamr-pipeline.html#glossary)

### Select the expected ASV and make mock_composition

You can now pick the correct sequences of the expected ASVs in each mock and make the [mock_composition](tutorial-vtamr-pipeline.html#mock_composition) file.

