---
title: "Short Tutorial vtamR"
author: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Short Tutorial vtamR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=TRUE,
  cache=FALSE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

# Short Tutorial

**Package version:** `r packageVersion("vtamR")`

## Documentation

This is a short version of the tutorial using the ZFZR demo data set.
For more details check out the appropriate section of the 
[detailed tutorial](tutorial-vtamr-pipeline.html)
and the documentation of the functions (`?function_name`)

> **Want to see the expected output without running the pipeline yourself?**
> Download a ready-made copy of this tutorial's output with:
> ```{r, eval=FALSE}
> vtamR::download_tutorial_output(dest = "my_folder/")
> ```
> This is useful for checking your own results against a reference, or for
> browsing the output structure before running the third-party tools
> (vsearch, blast, swarm, cutadapt) required for the pipeline itself.

## Set up

### Load library

```{r}
library(vtamR)
library(dplyr)
```


### Set paths (if needed)

If the third-party tools are not in your PATH, set the corresponding package options and specify the correct path to each tool.

See the [detailed tutorial on how to set the PATH](tutorial-vtamr-pipeline.html#set-path) for more information.

```{r, eval=FALSE}
options(vtamR.vsearch_path = "C:/Users/Public/vsearch-2.23.0-win-x86_64/bin/vsearch")
options(vtamR.blast_path = "blastn")
options(vtamR.cutadapt_path = "C:/Users/Public/cutadapt")
options(vtamR.swarm_path = "C:/Users/Public/swarm-3.1.5-win-x86_64/bin/swarm")
options(vtamR.pigz_path = "C:/Users/Public/pigz-win32/pigz")
```

```{r, echo = FALSE, eval=TRUE}
options(vtamR.vsearch_path = "~/miniconda3/envs/vtam/bin/vsearch")
options(vtamR.blast_path = "~/miniconda3/envs/vtam/bin/blastn")
options(vtamR.cutadapt_path = "~/miniconda3/envs/vtam_benchmark/bin/cutadapt")
options(vtamR.swarm_path = "swarm")
options(vtamR.pigz_path = "~/miniconda3/envs/vtam/bin/pigz")
```

## Input data
*See [Input data](tutorial-vtamr-pipeline.html#input-data) for more details.*

### Demo data, outdir, log file

```{r}
outdir <- "vtamR_demo_out/zfzr_plate1"
options(vtamR.log_file =  file.path(outdir, "vtamR_log.csv"))


fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <- system.file("extdata/demo/fastqinfo_zfzr_plate1.csv", package = "vtamR")
mock_ncbi_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package = "vtamR")
asv_list <- system.file("extdata/demo/ASV_list_with_IDs.csv", package = "vtamR") # all ASV of MFZR
```

### Database

```{r}
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")
```

## Pre-processing

To demonstrate different options, here, we will use a different order of the 
pre-processing steps

In the details tutorial we had
 - count_reads_batch
 - random_sample_batch
 - demultiplex_and_trim_fastq
 - merge_fastq_pairs
 
 Here we will place de subampling after the demmutliplex step:
 
 - demultiplex_and_trim_fastq
 - random_sample_batch_by
 - merge_fastq_pairs
 
### Demultiplex
*See [Demultiplex and Trim](tutorial-vtamr-pipeline.html#demultiplex) for more details.*

```{r}
demultiplexed_dir <- file.path(outdir, "1_demultiplexed")

fastqinfo_df <- demultiplex_and_trim_fastq(
  fastqinfo = fastqinfo,
  fastq_dir = fastq_dir,
  outdir = demultiplexed_dir,
  primer_to_end = FALSE,
  check_reverse = TRUE,
  compress = FALSE
)
```

### Subsampling
*See [Randomly Select Sequences](tutorial-vtamr-pipeline.html#subsample) for more details.*

If we used `random_sample_batch` with the `n` option as in the [detailed tutoriel](tutorial-vtamr-pipeline.html#subsample),
we would sample `n` sequences from each file and distort the relative read counts 
among samples.
To avoid this problem we will use `random_sample_batch_by` and the `N` option.
This allow the select `N` sequences all together from the files of each `replicate` (befined by the `by` argument),
while keeping the relative proportion of the read count among the files of the same 
replicate.

```{r}
randomseq_dir <- file.path(outdir, "2_random_seq")

fastqinfo_df <- random_sample_batch_by(
  info = fastqinfo_df,
  by = "replicate",
  dir = demultiplexed_dir,
  outdir = randomseq_dir,
  N = 40000,
  compress = FALSE
  ) 

```

### Merge Fastq Pairs
*See [Merge Fastq Pairs and Quality Filter](tutorial-vtamr-pipeline.html#merge) for more details.*

```{r}
merged_dir <- file.path(outdir, "3_merged")

sampleinfo_df <- merge_fastq_pairs(
  fastqinfo = fastqinfo_df,
  fastq_dir = randomseq_dir,
  outdir = merged_dir,
  fastq_maxee = 1,
  fastq_maxns = 0,
  fastq_allowmergestagger = T
)
```


### Dereplicate
*See [Dereplicate](tutorial-vtamr-pipeline.html#dereplicate) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "1_before_filter.csv")
updated_asv_list <- file.path(outdir, "ASV_list_with_IDs.csv")

read_count_df <- dereplicate(
  sampleinfo = sampleinfo_df,
  dir = merged_dir,
  outfile = outfile,
  input_asv_list = asv_list,
  output_asv_list = updated_asv_list
)
```


## Filtering
*See [Filtering](tutorial-vtamr-pipeline.html#filtering) for more details.*

### Initialize Statistics
*See [Initialize Statistics](tutorial-vtamr-pipeline.html#init-stat) for more details.*

```{r}
stat_df <- data.frame(
  parameters = character(),
  asv_count = integer(),
  read_count = integer(),
  sample_count = integer(),
  sample_replicate_count = integer()
)

stat_df <- get_stat(read_count_df, stat_df, stage = "input", params = NA)
```

### Filtering ASVs

#### Denoising
*See [Denoising with SWARM](tutorial-vtamr-pipeline.html#denoise) for more details.*

```{r}
split_clusters <- TRUE
by_sample <- TRUE

outfile <- file.path(outdir, "4_filter", "2_swarm_denoise.csv")
read_count_df <- denoise_by_swarm(
  read_count = read_count_df,
  outfile = outfile,
  by_sample = by_sample,
  split_clusters = split_clusters
)

param <- paste("by_sample:", by_sample, "split_clusters:", split_clusters)
stat_df <- get_stat(read_count_df, stat_df, stage="swarm_denoise", params=param)
```

#### Filter ASVs Global
*See [Filter out ASVs with Low Total Read Count](tutorial-vtamr-pipeline.html#filter_asv_global) for more details.*

```{r}
global_read_count_cutoff = 2
outfile <- file.path(outdir, "4_filter", "3_filter_asv_global.csv")

read_count_df <- filter_asv_global(
  read_count = read_count_df, 
  outfile = outfile, 
  cutoff = global_read_count_cutoff
  )

param <- paste("global_read_count_cutoff:", global_read_count_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_asv_global", params=param)
```

#### Filter ASVs Read Length
*See [Filter ASVs Based on Read Length](tutorial-vtamr-pipeline.html#filter_indel) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "4_filter_indel.csv")

read_count_df <- filter_indel(
  read_count = read_count_df, 
  outfile = outfile
  )
stat_df <- get_stat(read_count_df, stat_df, stage="filter_indel", params=NA)
```

#### Filter ASVs Stop Codons
*See [Filter ASVs with Stop Codons](tutorial-vtamr-pipeline.html#filter_stop_codon) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "5_filter_stop_codon.csv")

read_count_df <- filter_stop_codon(
  read_count = read_count_df, 
  outfile = outfile
  )

stat_df <- get_stat(read_count_df, stat_df, stage="filter_stop_codon", params=NA)
```

#### Filter Contaminant ASVs
*See [Filter Potential Contaminant ASVs](tutorial-vtamr-pipeline.html#filter_contaminant) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "6_filter_contaminant.csv")
conta_file <- file.path(outdir, "4_filter", "potential_contaminants.csv")

read_count_df <- filter_contaminant(
  read_count = read_count_df,
  sampleinfo = sampleinfo_df,
  conta_file = conta_file,
  outfile = outfile
)

stat_df <- get_stat(read_count_df, stat_df, stage="filter_contaminant", params=NA)
```

#### Filter Chimeric ASVs
*See [Filter Chimeric ASVs](tutorial-vtamr-pipeline.html#filter_chimera) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "7_filter_chimera.csv")

read_count_df <- filter_chimera(
  read_count = read_count_df,
  outfile = outfile
)

stat_df <- get_stat(read_count_df, stat_df, stage="filter_chimera", params=NA)
```

### Filter Replicates

*See [Filter Aberrant Replicates](tutorial-vtamr-pipeline.html#filter_replicate) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "8_filter_replicate.csv")
renkonen_distance_quantile <- 0.9

read_count_df <- filter_replicate(
  read_count = read_count_df,
  renkonen_distance_quantile = renkonen_distance_quantile,
  outfile = outfile
)

param <- paste("renkonen_distance_quantile:", renkonen_distance_quantile)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_replicate", params=param)
```

### Filtering Occurrences

#### Create `mock_composition` File
*See [Create `mock_composition` File](tutorial-vtamr-pipeline.html#match_variants_to_mock_species) for more details.*

```{r}
outdir_mock <- file.path(outdir, "mock_composition")

mock_template <- match_variants_to_mock_species(
  read_count = read_count_df,
  fas = mock_ncbi_fasta,
  taxonomy = taxonomy,
  sampleinfo = sampleinfo_df,
  outdir = outdir_mock
)

mock_composition <- file.path(outdir_mock, "mock_composition_template_to_check.csv")
mock_composition_df <- read.csv(mock_composition)
```

**Note:**

The `mock_composition_template_to_check.csv` file contains only 5 of the 6 expected ASVs.
Since the NCBI reference sequences were correctly chosen, this indicates that the ZFZP primer pair did not amplify *Phoxinus phoxinus*.

Because this sequence is absent from the `mock_composition`, it will not be counted as a false negative when evaluating performance. This is an artefact, and it should be kept in mind that it is in fact a false negative.

See [Create `mock_composition` File](tutorial-vtamr-pipeline.html#match_variants_to_mock_species) for more details.


#### Filter Occurrences by Samples
*See [Filter Low-Abundance Occurrences within Samples ](tutorial-vtamr-pipeline.html#filter_occurrence_sample) for more details.*

**Suggest parameter value**
```{r}
outfile = file.path(outdir, "4_filter", "suggest_parameters", "sample_cutoff.csv")

suggest_sample_cutoff_df <- suggest_sample_cutoff(
  read_count = read_count_df,
  mock_composition = mock_composition,
  outfile = outfile
)

head(suggest_sample_cutoff_df)
```

**Apply filter**
```{r}
sample_cutoff <- 0.0001
outfile <- file.path(outdir, "4_filter", "9_filter_occurrence_sample.csv")

read_count_df <- filter_occurrence_sample(
  read_count = read_count_df, 
  cutoff = sample_cutoff, 
  outfile = outfile
  )

param <- paste("cutoff:", sample_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_occurrence_sample", params=param)
```

#### Filter Occurrences by Replicate
*See [Filter Occurrences by Replicate Consistency](tutorial-vtamr-pipeline.html#filter_min_replicate-1) for more details.*

```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "10_filter_min_replicate.csv")

read_count_df <- filter_min_replicate(
  read_count = read_count_df, 
  cutoff = min_replicate_number, 
  outfile = outfile
  )

param <- paste("cutoff:", min_replicate_number)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_min_replicate", params=param)
```


#### Filter Occurrences by Variant and Read count

*See [Filter Low-Abundance Occurrences (`filter_occurrence_variant` and `filter_occurrence_read_count`)](tutorial-vtamr-pipeline.html#filter_occurrence_variant) for more details.*

**Suggest parameter values**
```{r}
outdir_suggest = file.path(outdir, "4_filter", "suggest_parameters")

suggest_variant_readcount_cutoffs_df <- suggest_variant_readcount_cutoffs(
  read_count = read_count_df,
  mock_composition = mock_composition,
  sampleinfo = sampleinfo_df,
  outdir = outdir_suggest,
  min_replicate_number = 2
)

head(suggest_variant_readcount_cutoffs_df)
```

**Filter occurrences by variant**
```{r}
variant_cutoff = 0.001
outfile <- file.path(outdir, "4_filter", "11_filter_occurrence_variant.csv")

read_count_df_variant <- filter_occurrence_variant(
  read_count = read_count_df, 
  cutoff = variant_cutoff, 
  outfile = outfile
  )
param <- paste("variant_cutoff:", variant_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_occurrence_variant", params=param)
```

**Filter occurrences by read count**

```{r}
read_count_cutoff <- 10
outfile <- file.path(outdir, "4_filter", "12_filter_occurrence_read_count.csv")

read_count_df_rc <- filter_occurrence_read_count(
  read_count = read_count_df, 
  cutoff = read_count_cutoff, 
  outfile = outfile
  )
param <- paste("read_count_cutoff:", read_count_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_occurrence_read_count", params=param)
```

**Pool results**
```{r}
outfile <- file.path(outdir, "4_filter", "13_pool_filters.csv")

read_count_df <- pool_filters(
  read_count_df_rc, read_count_df_variant, 
  outfile=outfile
  )

rm(read_count_df_rc)
rm(read_count_df_variant)

stat_df <- get_stat(read_count_df, stat_df, stage="pool_filters", params=NA)
```


#### Filter Occurrences by Replicate Again

*See [Filter Occurrences by Replicate Consistency](tutorial-vtamr-pipeline.html#filter_min_replicate-2) for more details.*

```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "14_filter_min_replicate.csv")

read_count_df <- filter_min_replicate(
  read_count = read_count_df, 
  cutoff = min_replicate_number, 
  outfile = outfile
  )
param <- paste("min_replicate_number:", min_replicate_number)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_min_replicate", params=param)
```

## Final steps

### Pool replicates
*See [Pool replicates](tutorial-vtamr-pipeline.html#pool_replicates) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "15_pool_replicates.csv")

read_count_sample_df <- pool_replicates(
  read_count = read_count_df,
  outfile = outfile
)
```

### Performance Metrics
*See [Performance Metrics](tutorial-vtamr-pipeline.html#classify_control_occurrences) for more details.*

```{r}
false_negatives <- file.path(outdir, "4_filter", "false_negatives.csv")
performance_metrics <- file.path(outdir, "4_filter", "performance_metrics.csv")
known_occurrences <- file.path(outdir, "4_filter", "known_occurrences.csv")

results <- classify_control_occurrences(
  read_count = read_count_sample_df, 
  sampleinfo = sampleinfo_df, 
  mock_composition = mock_composition, 
  known_occurrences = known_occurrences, 
  false_negatives = false_negatives,
  performance_metrics = performance_metrics
  )

head(read.csv(performance_metrics))
```

**Note:**
Keep in mind that one expected sequence in the mock is not present in the `mock_composition`.
Therefore, the true number of false negatives is 1, and the sensitivity is:

TP / (TP + FN) = 5 / (5 + 1) = 0.833


### Print Summary Statistics
*See [Print Summary Statistics](tutorial-vtamr-pipeline.html#print_stats) for more details.*

```{r}
run_info_dir <- file.path(outdir, "run_info")
check_dir(run_info_dir)
write.csv(stat_df, file = file.path(run_info_dir, "summary.csv"))
```


## Taxonomic Assignment
*See [Taxonomic Assignment](tutorial-vtamr-pipeline.html#taxonomic-assignment) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "assign_taxonomy_ltg.csv")

asv_tax <- assign_taxonomy_ltg(
  asv = read_count_sample_df,
  taxonomy = taxonomy,
  blast_db = blast_db,
  outfile = outfile
)
```

## Clustering (mOTUs)
*See [Clustering (mOTUs)](tutorial-vtamr-pipeline.html#cluster_asv) for more details.*

```{r}
outfile <- file.path(outdir, "5_cluster", "16_mOTU_vsearch.csv")

read_count_samples_df_ClusterSize <- cluster_asv(
  read_count = read_count_sample_df,
  method = "swarm",
  swarm_d = 7,
  outfile = outfile
)
```

## Output

### ASV table
*See [Output table](tutorial-vtamr-pipeline.html#write_asv_table) for more details.*

```{r write_asv_table}
outfile=file.path(outdir, "motu_table_taxa.csv")

asv_table_df <- write_asv_table(
  read_count = read_count_sample_df,
  outfile = outfile,
  asv_tax = asv_tax,
  sampleinfo = sampleinfo_df
)
```

### Convert output to a phyloseq object
```{r make_phyloseq}
# Create a data frame containing sample metadata
samples <- sampleinfo_df %>%
  select(sample, sample_type, habitat) %>%
  distinct()

phy_object <- format_for_phyloseq(
  otu = read_count_sample_df,
  tax = asv_tax,
  samples = sampleinfo_df,
  outfile = file.path(outdir, "phyloseq", "phyloseq.rds")
)
```

### Convert output for the `vegan` package

```{r format_to_vegan}
vegan_dfs <- format_for_vegan(
  otu = read_count_sample_df,
  tax = asv_tax,
  sample_type = sampleinfo_df,
  rm_coltrol = TRUE,
  outfile_motu = file.path(outdir, "vegan", "motu_table_vegan.csv"), 
  outfile_taxa = file.path(outdir, "vegan", "tax_table_vegan.csv")
)

vegan_motu_df <- vegan_dfs[[1]]
vegan_tax_df <- vegan_dfs[[2]]
```

## Document your workflow

### Record R and package versions 
```{r versions}
package_versions <- file.path(run_info_dir,  "R_package_versions.csv")
collect_package_info(package_versions)
```

### MIEM 

**MIEM bioinformatics**
```{r miem_bioinformatics}
miem_dir <- file.path(run_info_dir, "MIEM")
tmp <- miem_bioinformatics(
  r_versions = package_versions,
  outfile = file.path(miem_dir, "MIEM_bioinfo.csv")
  )
```


**MIEM Sequencing Summary**
```{r miem_results}
info_files = c(
  fastqinfo,
  file.path(outdir, "1_demultiplexed/fastqinfo.csv"),
  file.path(outdir, "2_random_seq/fastqinfo.csv"),
  file.path(outdir, "3_merged/fastainfo.csv")
)
read_count_files = c(
  file.path(outdir, "4_filter/1_before_filter.csv"),
  file.path(outdir, "4_filter/2_swarm_denoise.csv"),
  file.path(outdir, "4_filter/14_filter_min_replicate.csv"),
  file.path(outdir, "4_filter/15_pool_replicates.csv"),
  file.path(outdir, "5_cluster/16_mOTU_vsearch.csv")
)
taxa = file.path(outdir, "4_filter/assign_taxonomy_ltg.csv")

miem_results(
  info_files = info_files, 
  fastq_dir = fastq_dir, 
  read_count_files = read_count_files, 
  taxa = taxa, 
  outdir = miem_dir, 
  mock_composition = mock_composition,
  )
```
