---
title: "Ultra Short Tutorial vtamR - v1.0.1"
auhor: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Short Tutorial vtamR - v1.0.1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=TRUE,
  cache=TRUE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```


# vtamR Pipeline — Cheat Sheet

## 1. Setup

```{r}
library(vtamR)
```

```{r}
outdir <- "vtamR_demo_out/mfzr_plate1"

fastq_dir <- system.file("extdata/demo/fastq", package="vtamR")
fastqinfo <- system.file("extdata/demo/fastqinfo_mfzr_plate1.csv", package="vtamR")
mock_ncbi_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package="vtamR")
asv_list <- system.file("extdata/demo/asv_list.csv", package="vtamR")

taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package="vtamR")
blast_db <- file.path(system.file("extdata/db_test", package="vtamR"), "COInr_reduced")
```

---

## 2. Pre-processing

```{r}
fastainfo_df <- merge_fastq_pairs(
  fastqinfo, 
  fastq_dir, 
  outdir=file.path(outdir,"1_merged")
  )

fastainfo_df <- subsample_fasta(
  fastainfo_df, 
  fasta_dir=file.path(outdir,"1_merged"),
  outdir=file.path(outdir,"2_random_seq"),
  n=40000
  )

sampleinfo_df <- demultiplex_and_trim(
  fastainfo_df,
  fasta_dir=file.path(outdir,"2_random_seq"),
  outdir=file.path(outdir,"3_demultiplexed"),
  check_reverse=TRUE,
  cutadapt_minimum_length=170,
  cutadapt_maximum_length=185
  )

read_count_df <- dereplicate(
  sampleinfo_df,
  dir=file.path(outdir,"3_demultiplexed"),
  outfile=file.path(outdir,"4_filter","1_before_filter.csv"),
  input_asv_list=asv_list,
  output_asv_list=file.path(outdir,"ASV_list_with_IDs.csv")
  )
```


## 3. Filtering ASVs

```{r}
read_count_df <- denoise_by_swarm(
  read_count_df,
  outfile=file.path(outdir,"4_filter","2_swarm.csv"),
  by_sample=TRUE, 
  split_clusters=TRUE
  )

read_count_df <- filter_asv_global(
  read_count_df, 
  cutoff=2,
  outfile=file.path(outdir,"4_filter","3_global.csv")
  )

read_count_df <- filter_indel(
  read_count_df,
  outfile=file.path(outdir,"4_filter","4_indel.csv")
  )

read_count_df <- filter_stop_codon(
  read_count_df, 
  outfile=file.path(outdir,"4_filter","5_stop.csv")
  )

read_count_df <- filter_contaminant(
  read_count_df, 
  sampleinfo_df,
  conta_file=file.path(outdir,"4_filter","conta.csv"),
  outfile=file.path(outdir,"4_filter","6_conta.csv"))

read_count_df <- filter_chimera(
  read_count_df,
  vsearch_path=vsearch_path,
  outfile=file.path(outdir,"4_filter","7_chimera.csv")
  )
```


## 4. Filter Replicates

```{r}
read_count_df <- filter_replicate(
  read_count_df,
  renkonen_distance_quantile=0.9,
  outfile=file.path(outdir,"4_filter","8_replicate.csv")
  )
```


## 5. Filtering Occurrences

### Make mock composition
```{r}
mock_template <- match_variants_to_mock_species(
  read_count_df,
  fas=mock_ncbi_fasta,
  taxonomy=taxonomy,
  sampleinfo=sampleinfo_df,
  outdir=file.path(outdir,"mock")
  )

mock_composition <- file.path(outdir,"mock","mock_composition_template_to_check.csv")
```

### Suggest Parametres and Filter Occurrences

#### Filter by Sample

```{r}

sample_cutoff_df <- suggest_sample_cutoff(
  read_count=read_count_df,
  mock_composition=mock_composition,
  outfile=file.path(outdir, "4_filter", "suggest_parameters", "sample_cutoff.csv")
)
head(sample_cutoff_df)

read_count_df <- filter_occurrence_sample(
  read_count_df, 
  cutoff=0.004,
  outfile=file.path(outdir,"4_filter","9_sample.csv")
  )

read_count_df <- filter_min_replicate(
  read_count_df, 
  outfile=file.path(outdir,"4_filter","10_minrep.csv")
  )
```

#### Filter by Variant and Read Count

```{r}
variant_readcount_cutoffs_df <- suggest_variant_readcount_cutoffs(
  read_count_df,
  mock_composition = mock_composition,
  sampleinfo = sampleinfo_df,
  outdir= file.path(outdir, "4_filter", "suggest_parameters"),
  min_replicate_number = 2
)

head(variant_readcount_cutoffs_df)


read_count_df_variant <- filter_occurrence_variant(
  read_count_df, 
  cutoff=0.001,
  outfile=file.path(outdir,"4_filter","11_variant.csv")
  )                                

read_count_df_rc <- filter_occurrence_read_count(
  read_count_df, 
  cutoff=10,
  outfile=file.path(outdir,"4_filter","12_readcount.csv")
  )
                                                 
read_count_df <- pool_filters(
  read_count_df_rc, 
  read_count_df_variant,
  outfile=file.path(outdir,"4_filter","13_pool.csv")
  )                    

rm(read_count_df_rc, read_count_df_variant)
```

## 6. Final cleaning

```{r}
read_count_df <- filter_min_replicate(
  read_count_df, 
  outfile=file.path(outdir,"4_filter","14_minrep2.csv")
  )

read_count_sample_df <- pool_replicates(
  read_count_df,
  outfile=file.path(outdir,"4_filter","15_poolrep.csv")
  )
```


## 7. Taxonomy

```{r}
asv_tax <- assign_taxonomy_ltg(
  read_count_sample_df,
  taxonomy=taxonomy,
  blast_db=blast_db,
  outfile=file.path(outdir,"4_filter","taxonomy.csv")
  )
```


## 8. Clustering (mOTUs)

```{r}
motu_df <- cluster_asv(
  read_count_sample_df,
  method="vsearch", 
  identity=0.97,
  outfile=file.path(outdir,"5_cluster","motu.csv")
  )
```


## 9. Final table

```{r}
asv_table_df <- write_asv_table(
  read_count_sample_df,
  outfile=file.path(outdir,"motu_table_taxa.csv"),
  asv_tax=asv_tax,
  sampleinfo=sampleinfo_df
  )
```


## Pipeline summary

**FASTQ → merge → subsample → demultiplex → dereplicate → filters → pool → taxonomy → mOTUs → table**


