---
title: "Short Tutorial vtamR - v1.0.2"
auhor: Emese Meglécz
date:  "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Short Tutorial vtamR - v1.0.2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<style>
/* Block console output (<pre>) scrolls horizontally AND vertically */
pre {
  white-space: pre !important;        /* preserve spacing, no wrapping */
  overflow-x: auto !important;        /* horizontal scroll */
  overflow-y: auto !important;        /* vertical scroll */
  display: block;
  max-width: 100%;
  max-height: 500px;                  /* adjust height limit if you want */
  box-sizing: border-box;
}

/* Ensure code inside <pre> also stays fixed and scrollable */
pre code {
  white-space: pre !important;
}

/* TABLES: prevent wrapping, allow horizontal AND vertical scroll */
table {
  display: block;
  overflow-x: auto !important;
  overflow-y: auto !important;
  white-space: nowrap;                /* no wrapping of table cells */
  max-width: 100%;
  max-height: 500px;                  /* adjust as you like */
  box-sizing: border-box;
}
</style>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  echo=TRUE,
  eval=TRUE,
  cache=TRUE,
  fig.width = 6,
  out.width = '100%',
  comment = "#>"
)
```

# Short Tutorial

This is a short version of the tutorial using the ZFZR demo data set.
For more details check out the appropriate section of the 
[detailed tutorial](tutorial-vtamr-pipeline.html)
and the documentation of the functions (`?function_name`)

## Set up

### Load library

```{r}
library(vtamR)
```

### Set paths (if needed)

I suppose all third party tools are in your PATH. If not, see the
[detailed tutorial on how to set the PATH](tutorial-vtamr-pipeline.html#set-path).

## Input data
*See [Input data](tutorial-vtamr-pipeline.html#input-data) for more details.*

### Demo data

```{r}
outdir <- "vtamR_demo_out/zfzr_plate1"

fastq_dir <- system.file("extdata/demo/fastq", package = "vtamR")
fastqinfo <- system.file("extdata/demo/fastqinfo_zfzr_plate1.csv", package = "vtamR")
mock_ncbi_fasta <- system.file("extdata/demo/mock_ncbi.fasta", package = "vtamR")
asv_list <- system.file("extdata/demo/ASV_list_with_IDs.csv", package = "vtamR") # all ASV of MFZR
```

### Database

```{r}
taxonomy <- system.file("extdata/db_test/taxonomy_reduced.tsv", package = "vtamR")
blast_db <- system.file("extdata/db_test", package = "vtamR")
blast_db <- file.path(blast_db, "COInr_reduced")
```

## Pre-processing

### Merge Fastq Pairs
*See [Merge Fastq Pairs and Quality Filter](tutorial-vtamr-pipeline.html#merge) for more details.*

```{r}
merged_dir <- file.path(outdir, "1_merged")

fastainfo_df <- merge_fastq_pairs(
  fastqinfo=fastqinfo,
  fastq_dir=fastq_dir,
  outdir=merged_dir,
  fastq_maxee=1,
  fastq_maxns=0,
  fastq_allowmergestagger=T
)
```

### Subsampling
*See [Randomly Select Sequences](tutorial-vtamr-pipeline.html#subsample) for more details.*

```{r}
randomseq_dir <- file.path(outdir, "2_random_seq")

fastainfo_df <- subsample_fasta(
  fastainfo = fastainfo_df,
  fasta_dir = merged_dir,
  outdir = randomseq_dir,
  n = 40000
)
```

### Demultiplex
*See [Demultiplex and Trim](tutorial-vtamr-pipeline.html#demultiplex) for more details.*

```{r}
demultiplexed_dir <- file.path(outdir, "3_demultiplexed")

sampleinfo_df <- demultiplex_and_trim(
  fastainfo=fastainfo_df,
  fasta_dir=randomseq_dir,
  outdir=demultiplexed_dir,
  check_reverse=TRUE,
  cutadapt_minimum_length = 150,
  cutadapt_maximum_length = 165
)
```

### Dereplicate
*See [Dereplicate](tutorial-vtamr-pipeline.html#dereplicate) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "1_before_filter.csv")
updated_asv_list <- file.path(outdir, "ASV_list_with_IDs.csv")

read_count_df <- dereplicate(
  sampleinfo_df,
  dir=demultiplexed_dir,
  outfile=outfile,
  input_asv_list=asv_list,
  output_asv_list=updated_asv_list
)
```


## Filtering
*See [Filtering](tutorial-vtamr-pipeline.html#filtering) for more details.*

### Initialize Statistics
*See [Initialize Statistics](tutorial-vtamr-pipeline.html#init-stat) for more details.*

```{r}
stat_df <- data.frame(
  parameters = character(),
  asv_count = integer(),
  read_count = integer(),
  sample_count = integer(),
  sample_replicate_count = integer()
)

stat_df <- get_stat(read_count_df, stat_df, stage = "input", params = NA)
```

### Filtering ASVs

#### Denoising
*See [Denoising with SWARM](tutorial-vtamr-pipeline.html#denoise) for more details.*

```{r}
split_clusters <- TRUE
by_sample <- TRUE

outfile <- file.path(outdir, "4_filter", "2_swarm_denoise.csv")
read_count_df <- denoise_by_swarm(
  read_count_df,
  outfile=outfile,
  by_sample=by_sample,
  split_clusters=split_clusters
)

param <- paste("by_sample:", by_sample, "split_clusters:", split_clusters)
stat_df <- get_stat(read_count_df, stat_df, stage="swarm_denoise", params=param)
```

#### Filter ASVs Global
*See [Filter out ASVs with Low Total Read Count](tutorial-vtamr-pipeline.html#filter_asv_global) for more details.*

```{r}
global_read_count_cutoff = 2
outfile <- file.path(outdir, "4_filter", "3_filter_asv_global.csv")

read_count_df <- filter_asv_global(
  read_count_df, 
  outfile=outfile, 
  cutoff=global_read_count_cutoff
  )

param <- paste("global_read_count_cutoff:", global_read_count_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_asv_global", params=param)
```

#### Filter ASVs Read Length
*See [Filter ASVs Based on Read Length](tutorial-vtamr-pipeline.html#filter_indel) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "4_filter_indel.csv")

read_count_df <- filter_indel(
  read_count_df, 
  outfile=outfile
  )
stat_df <- get_stat(read_count_df, stat_df, stage="filter_indel", params=NA)
```

#### Filter ASVs Stop Codons
*See [Filter ASVs with Stop Codons](tutorial-vtamr-pipeline.html#filter_stop_codon) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "5_filter_stop_codon.csv")

read_count_df <- filter_stop_codon(
  read_count_df, 
  outfile=outfile
  )

stat_df <- get_stat(read_count_df, stat_df, stage="filter_stop_codon", params=NA)
```

#### Filter Contaminant ASVs
*See [Filter Potential Contaminant ASVs](tutorial-vtamr-pipeline.html#filter_contaminant) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "6_filter_contaminant.csv")
conta_file <- file.path(outdir, "4_filter", "potential_contaminants.csv")

read_count_df <- filter_contaminant(
  read_count_df,
  sampleinfo=sampleinfo_df,
  conta_file=conta_file,
  outfile=outfile
)

stat_df <- get_stat(read_count_df, stat_df, stage="filter_contaminant", params=NA)
```

#### Filter Chimeric ASVs
*See [Filter Chimeric ASVs](tutorial-vtamr-pipeline.html#filter_chimera) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "7_filter_chimera.csv")

read_count_df <- filter_chimera(
  read_count_df,
  outfile=outfile
)

stat_df <- get_stat(read_count_df, stat_df, stage="filter_chimera", params=NA)
```

### Filter Replicates

*See [Filter Aberrant Replicates](tutorial-vtamr-pipeline.html#filter_replicate) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "8_filter_replicate.csv")
renkonen_distance_quantile <- 0.9

read_count_df <- filter_replicate(
  read_count_df,
  renkonen_distance_quantile = renkonen_distance_quantile,
  outfile = outfile
)

param <- paste("renkonen_distance_quantile:", renkonen_distance_quantile)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_replicate", params=param)
```

### Filtering Occurrences

#### Create `mock_composition` File
*See [Create `mock_composition` File](tutorial-vtamr-pipeline.html#match_variants_to_mock_species) for more details.*

```{r}
outdir_mock <- file.path(outdir, "mock_composition")

mock_template <- match_variants_to_mock_species(
  read_count=read_count_df,
  fas=mock_ncbi_fasta,
  taxonomy=taxonomy,
  sampleinfo = sampleinfo_df,
  outdir= outdir_mock
)

mock_composition <- file.path(outdir_mock, "mock_composition_template_to_check.csv")
mock_composition_df <- read.csv(mock_composition)
```

**Note:**

The `mock_composition_template_to_check.csv` file contains only 5 of the 6 expected ASVs.
Since the NCBI reference sequences were correctly chosen, this indicates that the ZFZP primer pair did not amplify *Phoxinus phoxinus*.

Because this sequence is absent from the `mock_composition`, it will not be counted as a false negative when evaluating performance. This is an artefact, and it should be kept in mind that it is in fact a false negative.

See [Create `mock_composition` File](tutorial-vtamr-pipeline.html#match_variants_to_mock_species) for more details.


#### Filter Occurrences by Samples
*See [Filter Low-Abundance Occurrences within Samples ](tutorial-vtamr-pipeline.html#filter_occurrence_sample) for more details.*

**Suggest parameter value**
```{r}
outfile = file.path(outdir, "4_filter", "suggest_parameters", "sample_cutoff.csv")

suggest_sample_cutoff_df <- suggest_sample_cutoff(
  read_count=read_count_df,
  mock_composition=mock_composition,
  outfile=outfile
)

head(suggest_sample_cutoff_df)
```

**Apply filter**
```{r}
sample_cutoff <- 0.0001
outfile <- file.path(outdir, "4_filter", "9_filter_occurrence_sample.csv")

read_count_df <- filter_occurrence_sample(
  read_count_df, 
  cutoff=sample_cutoff, 
  outfile=outfile)

param <- paste("cutoff:", sample_cutoff)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_occurrence_sample", params=param)
```

#### Filter Occurrences by Replicate
*See [Filter Occurrences by Replicate Consistency](tutorial-vtamr-pipeline.html#filter_min_replicate-1) for more details.*

```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "10_filter_min_replicate.csv")

read_count_df <- filter_min_replicate(
  read_count_df, 
  cutoff=min_replicate_number, 
  outfile=outfile
  )

param <- paste("cutoff:", min_replicate_number)
stat_df <- get_stat(read_count_df, stat_df, stage="filter_min_replicate", params=param)
```


#### Filter Occurrences by Variant and Read count

*See [Filter Low-Abundance Occurrences (`filter_occurrence_variant` and `filter_occurrence_read_count`)](tutorial-vtamr-pipeline.html#filter_occurrence_variant) for more details.*

**Suggest parameter values**
```{r}
outdir_suggest = file.path(outdir, "4_filter", "suggest_parameters")

suggest_variant_readcount_cutoffs_df <- suggest_variant_readcount_cutoffs(
  read_count_df,
  mock_composition = mock_composition,
  sampleinfo = sampleinfo_df,
  outdir= outdir_suggest,
  min_replicate_number=2
)

head(suggest_variant_readcount_cutoffs_df)
```

**Filter occurrences by variant**
```{r}
variant_cutoff = 0.001
outfile <- file.path(outdir, "4_filter", "11_filter_occurrence_variant.csv")

read_count_df_variant <- filter_occurrence_variant(read_count_df, cutoff=variant_cutoff, outfile=outfile)
```

**Filter occurrences by read count**

```{r}
read_count_cutoff <- 10
outfile <- file.path(outdir, "4_filter", "12_filter_occurrence_read_count.csv")

read_count_df_rc <- filter_occurrence_read_count(
  read_count_df, 
  cutoff=read_count_cutoff, 
  outfile=outfile
  )
```

**Pool results**
```{r}
outfile <- file.path(outdir, "4_filter", "13_pool_filters.csv")

read_count_df <- pool_filters(
  read_count_df_rc, 
  read_count_df_variant, 
  outfile=outfile
  )

rm(read_count_df_rc)
rm(read_count_df_variant)
```


#### Filter Occurrences by Replicate Again

*See [Filter Occurrences by Replicate Consistency](tutorial-vtamr-pipeline.html#filter_min_replicate-2) for more details.*

```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "4_filter", "14_filter_min_replicate.csv")

read_count_df <- filter_min_replicate(
  read_count_df, 
  cutoff=min_replicate_number, 
  outfile=outfile
  )
```

## Final steps

### Pool replicates
*See [Pool replicates](tutorial-vtamr-pipeline.html#pool_replicates) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "15_pool_replicates.csv")

read_count_sample_df <- pool_replicates(
  read_count_df,
  outfile = outfile
)
```

### Performance Metrics
*See [Performance Metrics](tutorial-vtamr-pipeline.html#classify_control_occurrences) for more details.*

```{r}
false_negatives <- file.path(outdir, "4_filter", "false_negatives.csv")
performance_metrics <- file.path(outdir, "4_filter", "performance_metrics.csv")
known_occurrences <- file.path(outdir, "4_filter", "known_occurrences.csv")

results <- classify_control_occurrences(
  read_count_sample_df, 
  sampleinfo=sampleinfo_df, 
  mock_composition=mock_composition, 
  known_occurrences=known_occurrences, 
  false_negatives=false_negatives,
  performance_metrics=performance_metrics
  )

head(read.csv(performance_metrics))
```

**Note:**
Keep in mind that one expected sequence in the mock is not present in the `mock_composition`.
Therefore, the true number of false negatives is 1, and the sensitivity is:

TP / (TP + FN) = 5 / (5 + 1) = 0.833


### Print Summary Statistics
*See [Print Summary Statistics](tutorial-vtamr-pipeline.html#print_stats) for more details.*

```{r}
write.csv(stat_df, file = file.path(outdir, "summary.csv"))
```


## Taxonomic Assignment
*See [Taxonomic Assignment](tutorial-vtamr-pipeline.html#taxonomic-assignment) for more details.*

```{r}
outfile <- file.path(outdir, "4_filter", "assign_taxonomy_ltg.csv")

asv_tax <- assign_taxonomy_ltg(
  asv=read_count_sample_df,
  taxonomy=taxonomy,
  blast_db=blast_db,
  outfile=outfile
)
```

## Clustering (mOTUs)
*See [Clustering (mOTUs)](tutorial-vtamr-pipeline.html#cluster_asv) for more details.*

```{r}
outfile <- file.path(outdir, "5_cluster", "16_mOTU_vsearch.csv")

read_count_samples_df_ClusterSize <- cluster_asv(
  read_count_sample_df,
  method = "swarm",
  swarm_d  = 7,
  outfile=outfile
)
```

## Output table
*See [Output table](tutorial-vtamr-pipeline.html#write_asv_table) for more details.*

```{r}
outfile=file.path(outdir, "motu_table_taxa.csv")

asv_table_df <- write_asv_table(
  read_count_sample_df,
  outfile=outfile,
  asv_tax=asv_tax,
  sampleinfo=sampleinfo_df
)
```
