---
title: "Benchmark vtamR"
auhor: Emese Meglecz
date: "2025-09-23"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

# VTAM by default params

This pipeline reconstituate the VTAM pipline, using default parameters
It gives a final asv_table that is fine to identify expected mock sequences

!!!! Info on filtering parameters for vtam benchmark:
/home/meglecz/vtam_benchmark_local/metafiles/params_filter_fish_mfzr.yml

# from dereplicated dataset till end of filter with default parameters

```{r}
library(vtamR)
library(dplyr)

### set up
cutadapt_path <- "~/miniconda3/envs/vtam/bin/cutadapt"
vsearch_path <- "~/miniconda3/envs/vtam/bin/vsearch"
blast_path <- "~/miniconda3/envs/vtam/bin/blastn"
swarm_path <- "swarm"
num_threads <- 8
sep <- ","
```


## set path to marker (MFZR or ZFZR)
```{r}

dir <- "~/vtamR/local/benchmark/fish/results_zfzr"
outdir <- file.path(dir, "filter")
sortedinfo <-  file.path(dir, "sorted", "sortedinfo.csv")
input <- file.path(dir, "1_before_filter.csv")
read_count_df <- read.csv(input)
```



```{r}
stat_df <- data.frame(parameters=character(),
                      asv_count=integer(),
                      read_count=integer(),
                      sample_count=integer(),
                      sample_replicate_count=integer())

stat_df <- GetStat(read_count_df, stat_df, stage="input", params=NA)
```

## delete singletons
```{r}
global_read_count_cutoff = 2
outfile <- file.path(outdir, "2_LFNglobalReadCount.csv")
read_count_df <- LFNglobalReadCount(read_count_df, 
                                    cutoff=global_read_count_cutoff,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="LFNglobalReadCount", params=global_read_count_cutoff)
```


## LFN by default
```{r}
### LFNvariant
lnf_variant_cutoff = 0.001
by_replicate = TRUE
outfile <- file.path(outdir, "3_LFNvariant.csv")
read_count_df_lnf_variant <- LFNvariant(read_count_df, 
                                        cutoff=lnf_variant_cutoff,
                                    outfile=outfile, 
                                    by_replicate=by_replicate)
stat_df <- GetStat(read_count_df_lnf_variant, stat_df, stage="LFNvariant", params=lnf_variant_cutoff)

### LFNreadCount
lfn_read_count_cutoff <- 10
outfile <- file.path(outdir, "4_LFNreadCount.csv")
read_count_df_lfn_read_count <- LFNreadCount(read_count_df, 
                                             cutoff=lfn_read_count_cutoff,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df_lfn_read_count, stat_df, stage="LFNreadCount", params=lfn_read_count_cutoff)

### LFNsampleReplicate
lfn_sample_replicate_cutoff <- 0.001
outfile <- file.path(outdir, "5_LFNsampleReplicate.csv")
read_count_df_lfn_sample_replicate <- LFNsampleReplicate(read_count_df, 
                                    cutoff=lfn_sample_replicate_cutoff,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df_lfn_sample_replicate, stat_df, stage="LFNsampleReplicate", params=lfn_sample_replicate_cutoff)

### Combine results
outfile <- file.path(outdir, "6_PoolFilters.csv")
read_count_df <- PoolFilters(read_count_df_lfn_read_count, 
                             read_count_df_lnf_variant,
                             read_count_df_lfn_sample_replicate,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="Combine results", params=NA)

```

## FilterMinReplicate
```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "7_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```

## FilterPCRerror
```{r}
pcr_error_var_prop <- 0.10
max_mismatch <- 1
outfile <- file.path(outdir, "8_FilterPCRerror.csv")
read_count_df <- FilterPCRerror(read_count_df, 
                                vsearch_path=vsearch_path, 
                                num_threads=num_threads,
                                pcr_error_var_prop=pcr_error_var_prop, 
                                max_mismatch=max_mismatch,
                                    outfile=outfile)

par <- paste(pcr_error_var_prop, max_mismatch, sep=";" )
stat_df <- GetStat(read_count_df, stat_df, stage="FilterPCRerror", params=par)
```


## FilterChimera
```{r}
abskew=16
by_sample = T
sample_prop = 0.8
outfile <- file.path(outdir, "9_FilterChimera.csv")
read_count_df <- FilterChimera(read_count_df, 
                               vsearch_path=vsearch_path, 
                               num_threads=num_threads,
                               by_sample=by_sample, 
                               sample_prop=sample_prop, 
                               abskew=abskew,
                                    outfile=outfile)
par <- paste(abskew, by_sample, sample_prop, sep=";" )
stat_df <- GetStat(read_count_df, stat_df, stage="FilterChimera", params=par)
```

## FilterMinReplicate
```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "10_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```

## FilterRenkonen
```{r}
#### FilterRenkonen
renkonen_within_df <- MakeRenkonenDistances(read_count_df, compare_all=FALSE)

DensityPlot_Renkonen <- DensityPlot_RenkonenDistance(renkonen_within_df)
Barplot_Renkonen <- Barplot_RenkonenDistance(renkonen_within_df, sample_types=sortedinfo, x_axis_label_size=6)

# in VTAM by default it is the 9th of the value that are under the threshlod => get this value from graphic
# ZFZR
cutoff <- 0.17793
# MFZR
cutoff <- 0.198999

renkonen_distance_quantile=0.9
outfile <- file.path(outdir, "11_FilterRenkonen.csv")
read_count_df <- FilterRenkonen(read_count_df, 
                                renkonen_distance_quantile=renkonen_distance_quantile,
                                    outfile=outfile)


stat_df <- GetStat(read_count_df, stat_df, stage="Renkonen", params=renkonen_distance_quantile)
```

## FilterMinReplicate
```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "12_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```

## FilterIndel
```{r}
outfile <- file.path(outdir, "13_FilterIndel.csv")
read_count_df <- FilterIndel(read_count_df,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterIndel", params=NA)
```


## FilterCodonStop
```{r}
genetic_code = 5
outfile <- file.path(outdir, "14_FilterCodonStop.csv")
read_count_df <- FilterCodonStop(read_count_df, 
                                 genetic_code=genetic_code,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterCodonStop", params=NA)
```



# make mock_composition

# pool replicates and taxassign
```{r}
outfile <- file.path(outdir, "15_PoolReplicates.csv")
read_count_samples_df <- PoolReplicates(read_count_df,
                                    outfile=outfile)
stat_df <- GetStat(read_count_samples_df, 
                   stat_df, 
                   stage="PoolReplicates"
                   )


taxonomy  <- "/home/meglecz/mkCOInr/COInr/COInr_for_vtam_2025_05_23_dbV5/COInr_for_vtam_taxonomy.tsv"
blast_db <- "/home/meglecz/mkCOInr/COInr/COInr_for_vtam_2025_05_23_dbV5/COInr_for_vtam"
blast_path <- "~/miniconda3/envs/vtam/bin/blastn"

outfile <- file.path(outdir, "16_TaxAssign.csv")
asv_tax <- TaxAssign(asv=read_count_samples_df, 
                     taxonomy=taxonomy, 
                     blast_db=blast_db, 
                     blast_path=blast_path, 
                     outfile=outfile, 
                     num_threads=num_threads
                     )


outfile=file.path(outdir, "Asvtable_with_TaxAssign_prefilter.csv")
asv_table_df <- WriteASVtable(read_count_samples_df, 
                              outfile=outfile, 
                              asv_tax=asv_tax, 
                              sortedinfo=sortedinfo, 
                              add_empty_samples=T, 
                              add_sums_by_sample=T, 
                              add_sums_by_asv=T, 
                              )
```


Mock composition

- Caenis pusilla => OK
- Baetis rhodani => OK
- Orthocladiinae => Rheocricotopus chalybeatus
- Chironomidae => Synorthocladius semivirens => tolerate
- Hydropsyche pellucidula => OK
- Phoxinus phoxinus => OK => tolerate

# Make mock_composition by hand ~/vtamR/local/benchmark/fish/data/mock_composition_###.csv

# write stat
```{r}
stat_file <- file.path(outdir, "stat_file.csv")
write.csv(stat_df, stat_file, row.names = TRUE)
```