---
title: "Benchmark vtamR"
auhor: Emese Meglecz
date: "2025-09-23"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

!!!! Info on filtering parameters for vtam benchmark:
/home/meglecz/vtam_benchmark_local/metafiles/params_filter_fish_mfzr.yml

# VTAM optimized
Follows the VTAM pipeline getting and using optimized parametres.
Needs a mock_composition file

```{r}
library(vtamR)
library(dplyr)

### set up
cutadapt_path <- "~/miniconda3/envs/vtam/bin/cutadapt"
vsearch_path <- "~/miniconda3/envs/vtam/bin/vsearch"
blast_path <- "~/miniconda3/envs/vtam/bin/blastn"
swarm_path <- "swarm"
num_threads <- 8
sep <- ","
```


## set path to marker (MFZR or ZFZR)
```{r}

dir <- "~/vtamR/local/benchmark/fish/results_zfzr"
mock_composition <- "/home/meglecz/vtamR/local/benchmark/fish/data/mock_composition_zfzr.csv"
outdir <- file.path(dir, "filter_optimised")
sortedinfo <-  file.path(dir, "sorted", "sortedinfo.csv")
input <- file.path(dir, "1_before_filter.csv")
read_count_df <- read.csv(input)
```



```{r}
stat_df <- data.frame(parameters=character(),
                      asv_count=integer(),
                      read_count=integer(),
                      sample_count=integer(),
                      sample_replicate_count=integer())

stat_df <- GetStat(read_count_df, stat_df, stage="input", params=NA)
```

## delete singletons
```{r}
global_read_count_cutoff = 2
outfile <- file.path(outdir, "2_LFNglobalReadCount.csv")
read_count_df <- LFNglobalReadCount(read_count_df, 
                                    cutoff=global_read_count_cutoff,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="LFNglobalReadCount", params=global_read_count_cutoff)
```


## OptimizeLFNsampleReplicate
```{r}
outfile = file.path(outdir, "OptimizeLFNsampleReplicate.csv")
OptimizeLFNsampleReplicate_df <- OptimizeLFNsampleReplicate(read_count=read_count_df,
                                                            mock_composition=mock_composition,
                                                            outfile=outfile)
# ZFZR 
# Hydropsyche pellucidula < 10 reads
# Phoxinus phoxinus 0 read
# make them tolerate and not keep

### LFNsampleReplicate
# MFZR 0.003
# ZFZR 0.001
lfn_sample_replicate_cutoff <- 0.001
outfile <- file.path(outdir, "3_LFNsampleReplicate.csv")
read_count_df <- LFNsampleReplicate(read_count_df, 
                                    cutoff=lfn_sample_replicate_cutoff,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="LFNsampleReplicate", params=lfn_sample_replicate_cutoff)


min_replicate_number <- 2
outfile <- file.path(outdir, "4_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```


## OptimizePCRerror
```{r}

outfile <- file.path(outdir, "OptimizePCRerror.csv")
OptimizePCRerror_df <- OptimizePCRerror(read_count_df, 
                                        mock_composition=mock_composition, 
                                        vsearch_path=vsearch_path, 
                                        outfile=outfile,
                                        min_read_count=5
                                        )
# MFZR ZFZR
pcr_error_var_prop <- 0.10
max_mismatch <- 1
outfile <- file.path(outdir, "5_FilterPCRerror.csv")
read_count_df <- FilterPCRerror(read_count_df, 
                                vsearch_path=vsearch_path, 
                                num_threads=num_threads,
                                pcr_error_var_prop=pcr_error_var_prop, 
                                max_mismatch=max_mismatch,
                                    outfile=outfile)

par <- paste(pcr_error_var_prop, max_mismatch, sep=";" )
stat_df <- GetStat(read_count_df, stat_df, stage="FilterPCRerror", params=par)


min_replicate_number <- 2
outfile <- file.path(outdir, "6_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)

```




## OptimizeLFNreadCountLFNvariant_df
```{r}
# Detect known occurrences
results <- MakeKnownOccurrences(read_count = read_count_df, 
                                sortedinfo=sortedinfo, 
                                mock_composition=mock_composition
                                )

# give explicit names to the 3 output data frames
known_occurrences_df <- results[[1]]
missing_occurrences_df <- results[[2]]
performance_metrics_df <- results[[3]]


# OptimizeLFNreadCountLFNvariant_df
outfile = file.path(outdir, "OptimizeLFNreadCountLFNvariant.csv")
by_replicate = TRUE
OptimizeLFNreadCountLFNvariant_df <- OptimizeLFNreadCountLFNvariant(
  read_count_df,
  known_occurrences=known_occurrences_df,
  outfile= outfile, 
  min_replicate_number=2,
  by_replicate=by_replicate
  )


### LFNvariant
# MFZR 0.004 suggested by vtamR, but 0.006 used in benchmark, probably, because in VTAM the complete dataset was used; difficult to reproduce exactly the VTAM algo, use the value suggested by OptimizeLFNreadCountLFNvariant
lnf_variant_cutoff = 0.004
# ZFZR
lnf_variant_cutoff = 0.001
outfile <- file.path(outdir, "7_LFNvariant.csv")
by_replicate=TRUE
read_count_df_lnf_variant <- LFNvariant(read_count_df, 
                                        cutoff=lnf_variant_cutoff, 
                                        outfile=outfile,
                                        by_replicate=by_replicate
                                        )
par <- paste(lnf_variant_cutoff, by_replicate, sep=";")
stat_df <- GetStat(read_count_df_lnf_variant, 
                   stat_df, 
                   stage="LFNvariant", 
                   params=par)

# MFZR
lfn_read_count_cutoff <- 10
# ZFZR
lfn_read_count_cutoff <- 35
outfile <- file.path(outdir, "8_LFNreadCount.csv")
read_count_df_lfn_read_count <- LFNreadCount(read_count_df, 
                                             cutoff=lfn_read_count_cutoff, 
                                             outfile=outfile
                                             )
stat_df <- GetStat(read_count_df_lfn_read_count, 
                   stat_df, stage="LFNreadCount", 
                   params=lfn_read_count_cutoff
                   )


outfile <- file.path(outdir, "9_poolLFN.csv")
read_count_df <- PoolFilters(read_count_df_lfn_read_count, 
                             read_count_df_lnf_variant, 
                             outfile=outfile
                             )
stat_df <- GetStat(read_count_df, 
                   stat_df, 
                   stage="FilterLFN"
                   )
# delete temporary data frames
rm(read_count_df_lfn_read_count)
rm(read_count_df_lnf_variant)


min_replicate_number <- 2
outfile <- file.path(outdir, "10_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)

```


## FilterChimera
```{r}
abskew=16
by_sample = T
sample_prop = 0.8
outfile <- file.path(outdir, "11_FilterChimera.csv")
read_count_df <- FilterChimera(read_count_df, 
                               vsearch_path=vsearch_path, 
                               num_threads=num_threads,
                               by_sample=by_sample, 
                               sample_prop=sample_prop, 
                               abskew=abskew,
                                    outfile=outfile)
par <- paste(abskew, by_sample, sample_prop, sep=";" )
stat_df <- GetStat(read_count_df, stat_df, stage="FilterChimera", params=par)
```

## FilterMinReplicate
```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "12_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```

## FilterRenkonen
```{r}
#### FilterRenkonen
renkonen_within_df <- MakeRenkonenDistances(read_count_df, compare_all=FALSE)

DensityPlot_Renkonen <- DensityPlot_RenkonenDistance(renkonen_within_df)
Barplot_Renkonen <- Barplot_RenkonenDistance(renkonen_within_df, sample_types=sortedinfo, x_axis_label_size=6)

# MFZR cutoff 0.188186
# ZFZR cutoff 0.151292
renkonen_distance_quantile=0.9
outfile <- file.path(outdir, "13_FilterRenkonen.csv")
read_count_df <- FilterRenkonen(read_count_df, 
                                renkonen_distance_quantile=renkonen_distance_quantile,
                                    outfile=outfile)


stat_df <- GetStat(read_count_df, stat_df, stage="Renkonen", params=renkonen_distance_quantile)
```

## FilterMinReplicate
```{r}
min_replicate_number <- 2
outfile <- file.path(outdir, "14_FilterMinReplicate.csv")
read_count_df <- FilterMinReplicate(read_count_df, 
                                    cutoff=min_replicate_number,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterMinReplicate", params=min_replicate_number)
```

## FilterIndel
```{r}
outfile <- file.path(outdir, "15_FilterIndel.csv")
read_count_df <- FilterIndel(read_count_df,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterIndel", params=NA)
```


## FilterCodonStop
```{r}
genetic_code = 5
outfile <- file.path(outdir, "16_FilterCodonStop.csv")
read_count_df <- FilterCodonStop(read_count_df, 
                                 genetic_code=genetic_code,
                                    outfile=outfile)
stat_df <- GetStat(read_count_df, stat_df, stage="FilterCodonStop", params=NA)
```


## PoolReplicates
```{r}
outfile <- file.path(outdir, "17_PoolReplicates.csv")
read_count_samples_df <- PoolReplicates(read_count_df, 
                                        outfile=outfile
                                        )
stat_df <- GetStat(read_count_samples_df, 
                   stat_df, 
                   stage="PoolReplicates"
                   )
```


## Get performance_metrics
```{r}
missing_occurrences <- file.path(outdir, "missing_occurrences.csv")
performance_metrics <- file.path(outdir, "performance_metrics.csv")
known_occurrences <- file.path(outdir, "known_occurrences.csv")

results <- MakeKnownOccurrences(read_count_samples_df, 
                                sortedinfo=sortedinfo, 
                                mock_composition=mock_composition, 
                                known_occurrences=known_occurrences, 
                                missing_occurrences=missing_occurrences,
                                performance_metrics=performance_metrics
                                )
# give explicit names to the 3 output data frames
known_occurrences_df <- results[[1]]
missing_occurrences_df <- results[[2]]
performance_metrics_df <- results[[3]]
```


# write stat
```{r}
stat_file <- file.path(outdir, "stat_file.csv")
write.csv(stat_df, stat_file, row.names = TRUE)
```